### 2024-01-24 manuscript/manuscript.Rmd
trying out Rmd rticle for this one

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-apa      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: "Arriaga meets Kitagawa. life expectancy decompositions including population subgroups"

titlerunning: Arriaga meets Kitagawa

authors: 
  - prefix: Dr.
    firstname: Timothy
    #particle: van 
    lastname: Riffe
    #suffix: III
    #naturename: Poet Laureate
    #degrees: MSc, PhD
    email: tim.riffe@ehu.eus
    affiliation: [1,2,3]
    corresponding: TRUE
#    equalcont: These authors contributed equally to this work.

  # - firstname: Second
  #   lastname: Author
  #   affiliation: 2
  #   equalcont: These authors contributed equally to this work.

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: Department of Sociology and Social Work
      orgname: University of the Basque Country (UPV/EHU)
    address:
        street: Barrio Sarriena s/n
        city: Leioa
        postcode: 48940
        state: Bizkaia
        country: Spain
  - number: 2
    corresponding: FALSE
    info:
      orgname: Ikerbasque (Basque Foundation for Science)
  - number: 3
    corresponding: FALSE
    info:
      orgdiv: Laboratory of Population Health
      orgname: Max Planck Institute for Demographic Research

keywords:
  - Decomposition
  - Mortality
  - Cause of death
  - Population Structure
  - Mortality Inequalities

# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

# Sample for unstructured abstract 
# abstract: |
#   The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.

# Sample for structured abstract
abstract: |
 **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age. A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into effects from differences in structure and differences in the weighted value.

 **Objective:** Sometimes we would like to decompose a difference between two populations that are each composed of like-defined subpopulations. Said decomposition would produce effects for each age of each subpopulation, as well as a marginal effect of composition differences. I propose a straightforward analytic method to do this.

 **Methods:** In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation.

 **Results:** I show the proposed analytic method to be equivalent to a Horiuchi et al (2008) reframing of the same problem. Some mentionable properties: (i) There is no limit to the number of subpopulations, (ii) it is straightforward to incorporate cause-of-death information, (iii) composition is here only considered in the radix age. I currently have results for simulated mortality rates, but I promise to wrangle up an empirical application to demonstrate the method.

 **Conclusions:** The analytic decomposition method I propose is advantageous compared to a Horiuchi method for this problem purely for reasons of computational efficiency. This method could help further disentangle the effects of mortality and composition differences in explaining or clarifying paradoxes or secular change. I promise to think further about subpopulation weighting that might occur over all ages (using prevalence information).}

bibliography: bibliography.bib

header-includes: |
  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

output: rticles::springer_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")  # This forces knitr to label all figures.
```

# Introduction

An @arriaga1984measuring decomposition allows us to decompose changes in life expectancy into contributions due to mortality changes in different ages. The method was designed to be practical, and framed in terms of age lifetable columns expressed in discrete age groups. A well-known property of the method is that mortality changes in different ages need not be proportional. Derived contributions also sum exactly to the observed life expectancy difference. A not-well-known property of the method is that it is asymmetrical, in the sense that the absolute values of age-specific contributions depend on whether we compare population 1 with 2, or population 2 with 1. A further property of the method is that it is designed to work with homogeneous populations, meaning that populations 1 and 2 are each homogeneous, in the sense that each is expressed with only one lifetable.

A @kitagawa1955components decomposition allows us to decompose differences in a weighted mean due to differences in weights (structure) and differences in the value being weighted (often rates, but in our case life expectancies). This widely-used decomposition method is well-known to be exact in that the resulting structure and value components sum exactly to the observed difference in weighted means. The individual elements (ages, or life expectancies for us) of the value being weighted have identifiable effects. It is not well-known that the individual elements of the structure component do not have identifiable effects. Rather, the structure effects should be summed to a a marginal effect due to differences in structure.

Sometimes we have a situation where we would like to decompose a difference between two populations that are each composed of like-defined subpopulations. For example, life expectancy in France versus Spain, each with education-specific subpopulations. In such situations, a decomposition should tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change. In this paper, I would like to propose a straightforward analytic method to decompose in this way. In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation so as to isolate the age-subpopulation-specific effects on the overall life expectancy difference. I will justify this rescaling, and show that the results of this procedure are fully consistent with the results of a @horiuchi2008decomposition reframing of the problem.

# Method

## Notation

Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et. al. (2024), from which we here consider only symmetrical Arriaga decompositions. This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results. For this, we use the following lifetable columns:

```{=tex}
\begin{itemize}

\item{$\ell(a)$} lifetable survivorship at exact age $a$.

\item{${}_nL_a$} lifetable person years lived in the interval $[a,a+n)$.

\item{$T_a$} total lifetable person years lived beyond age $a$.

\item{$e_a$} remaining life expectancy at exact age $a$.

\end{itemize}
```
We also use the superscript $s$ to index subpopulations comprising the total population, and the superscript $t$ to index time points, or some other way of differentiating the total population. The index $e^{s,t}_0$ reads as "life expectancy at birth for subpopulation $s$ at time $t$", e.g. $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.

Then one can write Arriaga's standard decomposition method like as per eq~\eqref{eq_arriaga1}. To keep things simple, we assume a lifetable radix of 1, meaning $\ell_0 = 1$.

$$
\begin{aligned}
{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
 &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
\end{aligned}
$$

\backmatter

\bmhead{Supplementary information}

If your article has accompanying supplementary file/s please state so here.

Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information. This may be requested by the editorial team/s if it is missing.

Please refer to Journal-level guidance for any specific requirements.

\bmhead{Acknowledgments}

Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

# Declarations {.unnumbered}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading \`Declarations':

-   Funding
-   Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-   Ethics approval
-   Consent to participate
-   Consent for publication
-   Availability of data and materials
-   Code availability
-   Authors' contributions

\noindent If any of the sections are not relevant to your manuscript, please include the heading and write \`Not applicable' for that section.

::: {.flushleft latex="true"}
Editorial Policies for:

\noindent Springer journals and proceedings: <https://www.springer.com/gp/editorial-policies>

\noindent Nature Portfolio journals: <https://www.nature.com/nature-research/editorial-policies>

\noindent \textit{Scientific Reports}: <https://www.nature.com/srep/journal-policies/editorial-policies>

\noindent BMC journals: <https://www.biomedcentral.com/getpublished/editorial-policies>
:::

::: {.appendices latex="true"}
# Section title of first appendix {#secA1}

An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.

For submissions to Nature Portfolio Journals please use the heading "Extended Data".
:::

### 2024-01-24 manuscript/manuscript.tex
trying out Rmd rticle for this one

--- 2024-01-24 manuscript/manuscript.Rmd
@@
----
-# Supported options: 
-#   sn-nature:       Style for submissions to Nature Portfolio journals
-#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
-#   sn-mathphys:     Math and Physical Sciences Reference Style
-#   sn-aps:          American Physical Society (APS) Reference Style
-#   sn-vancouver:    Vancouver Reference Style
-#   sn-apa:          APA Reference Style 
-#   sn-chicago:      Chicago-based Humanities Reference Style
-#   default:         Default
+%Version 2.1 April 2023
+% See section 11 of the User Manual for version history
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-classoptions: 
-  - sn-apa      
-  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
-  # - referee       # Optional: Use double line spacing 
-  # - lineno        # Optional: Add line numbers
-  # - iicol         # Optional: Double column layour
+\documentclass[sn-apa,pdflatex]{sn-jnl}
 
-title: "Arriaga meets Kitagawa. life expectancy decompositions including population subgroups"
+%%%% Standard Packages
+%%<additional latex packages if required can be included here>
 
-titlerunning: Arriaga meets Kitagawa
+\usepackage{graphicx}%
+\usepackage{multirow}%
+\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{amsthm}%
+\usepackage{mathrsfs}%
+\usepackage[title]{appendix}%
+\usepackage{xcolor}%
+\usepackage{textcomp}%
+\usepackage{manyfoot}%
+\usepackage{booktabs}%
+\usepackage{algorithm}%
+\usepackage{algorithmicx}%
+\usepackage{algpseudocode}%
+\usepackage{listings}%
+%%%%
 
-authors: 
-  - prefix: Dr.
-    firstname: Timothy
-    #particle: van 
-    lastname: Riffe
-    #suffix: III
-    #naturename: Poet Laureate
-    #degrees: MSc, PhD
-    email: tim.riffe@ehu.eus
-    affiliation: [1,2,3]
-    corresponding: TRUE
-#    equalcont: These authors contributed equally to this work.
+%%%%%=============================================================================%%%%
+%%%%  Remarks: This template is provided to aid authors with the preparation
+%%%%  of original research articles intended for submission to journals published
+%%%%  by Springer Nature. The guidance has been prepared in partnership with
+%%%%  production teams to conform to Springer Nature technical requirements.
+%%%%  Editorial and presentation requirements differ among journal portfolios and
+%%%%  research disciplines. You may find sections in this template are irrelevant
+%%%%  to your work and are empowered to omit any such section if allowed by the
+%%%%  journal you intend to submit to. The submission guidelines and policies
+%%%%  of the journal take precedence. A detailed User Manual is available in the
+%%%%  template package for technical guidance.
+%%%%%=============================================================================%%%%
 
-  # - firstname: Second
-  #   lastname: Author
-  #   affiliation: 2
-  #   equalcont: These authors contributed equally to this work.
-
-affiliations:
-  - number: 1
-    corresponding: TRUE
-    info:
-      orgdiv: Department of Sociology and Social Work
-      orgname: University of the Basque Country (UPV/EHU)
-    address:
-        street: Barrio Sarriena s/n
-        city: Leioa
-        postcode: 48940
-        state: Bizkaia
-        country: Spain
-  - number: 2
-    corresponding: FALSE
-    info:
-      orgname: Ikerbasque (Basque Foundation for Science)
-  - number: 3
-    corresponding: FALSE
-    info:
-      orgdiv: Laboratory of Population Health
-      orgname: Max Planck Institute for Demographic Research
-
-keywords:
-  - Decomposition
-  - Mortality
-  - Cause of death
-  - Population Structure
-  - Mortality Inequalities
-
-# pacs: 
-#   jel:
-#     - "D8"
-#     - "H51"
-#   msc:
-#     - "35A01"
-#     - "65L10"
-
-# Sample for unstructured abstract 
-# abstract: |
-#   The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications.
-Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.
-
-# Sample for structured abstract
-abstract: |
- **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age.
-A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into effects from differences in structure and differences in the weighted value.
-
- **Objective:** Sometimes we would like to decompose a difference between two populations that are each composed of like-defined subpopulations.
-Said decomposition would produce effects for each age of each subpopulation, as well as a marginal effect of composition differences.
-I propose a straightforward analytic method to do this.
-
- **Methods:** In short, within-subpopulation life expectancy differences can be handled by the Arriaga method.
-The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method.
-The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation.
-
- **Results:** I show the proposed analytic method to be equivalent to a Horiuchi et al (2008) reframing of the same problem.
-Some mentionable properties: (i) There is no limit to the number of subpopulations, (ii) it is straightforward to incorporate cause-of-death information, (iii) composition is here only considered in the radix age.
-I currently have results for simulated mortality rates, but I promise to wrangle up an empirical application to demonstrate the method.
-
- **Conclusions:** The analytic decomposition method I propose is advantageous compared to a Horiuchi method for this problem purely for reasons of computational efficiency.
-This method could help further disentangle the effects of mortality and composition differences in explaining or clarifying paradoxes or secular change.
-I promise to think further about subpopulation weighting that might occur over all ages (using prevalence information).}
-
-bibliography: bibliography.bib
-
-header-includes: |
 %% Per the spinger doc, new theorem styles can be included using built in style, 
@@
 
-output: rticles::springer_article
----
+\raggedbottom
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE,
-                      out.extra = "")  # This forces knitr to label all figures.
-```
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-# Introduction
+\begin{document}
 
-An @arriaga1984measuring decomposition allows us to decompose changes in life expectancy into contributions due to mortality changes in different ages.
-The method was designed to be practical, and framed in terms of age lifetable columns expressed in discrete age groups.
-A well-known property of the method is that mortality changes in different ages need not be proportional.
-Derived contributions also sum exactly to the observed life expectancy difference.
-A not-well-known property of the method is that it is asymmetrical, in the sense that the absolute values of age-specific contributions depend on whether we compare population 1 with 2, or population 2 with 1.
-A further property of the method is that it is designed to work with homogeneous populations, meaning that populations 1 and 2 are each homogeneous, in the sense that each is expressed with only one lifetable.
+\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. life expectancy
+decompositions including population subgroups}
 
-A @kitagawa1955components decomposition allows us to decompose differences in a weighted mean due to differences in weights (structure) and differences in the value being weighted (often rates, but in our case life expectancies).
-This widely-used decomposition method is well-known to be exact in that the resulting structure and value components sum exactly to the observed difference in weighted means.
-The individual elements (ages, or life expectancies for us) of the value being weighted have identifiable effects.
-It is not well-known that the individual elements of the structure component do not have identifiable effects.
-Rather, the structure effects should be summed to a a marginal effect due to differences in structure.
+%%=============================================================%%
+%% Prefix	-> \pfx{Dr}
+%% GivenName	-> \fnm{Joergen W.}
+%% Particle	-> \spfx{van der} -> surname prefix
+%% FamilyName	-> \sur{Ploeg}
+%% Suffix	-> \sfx{IV}
+%% NatureName	-> \tanm{Poet Laureate} -> Title after name
+%% Degrees	-> \dgr{MSc, PhD}
+%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
+%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
+%%=============================================================%%
 
-Sometimes we have a situation where we would like to decompose a difference between two populations that are each composed of like-defined subpopulations.
-For example, life expectancy in France versus Spain, each with education-specific subpopulations.
-In such situations, a decomposition should tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change.
-In this paper, I would like to propose a straightforward analytic method to decompose in this way.
-In short, within-subpopulation life expectancy differences can be handled by the Arriaga method.
-The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method.
-The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation so as to isolate the age-subpopulation-specific effects on the overall life expectancy difference.
-I will justify this rescaling, and show that the results of this procedure are fully consistent with the results of a @horiuchi2008decomposition reframing of the problem.
+\author*[1,2,3]{\pfx{Dr.} \fnm{Timothy} \sur{Riffe} }\email{\href{mailto:tim.riffe@ehu.eus}{\nolinkurl{tim.riffe@ehu.eus}}}
 
-# Method
+  \affil*[1]{\orgdiv{Department of Sociology and Social
+Work}, \orgname{University of the Basque Country
+(UPV/EHU)}, \orgaddress{\city{Leioa}, \country{Spain}, \postcode{48940}, \state{Bizkaia}, \street{Barrio
+Sarriena s/n}}}
+  \affil[2]{\orgname{Ikerbasque (Basque Foundation for Science)}}
+  \affil[3]{\orgdiv{Laboratory of Population Health}, \orgname{Max
+Planck Institute for Demographic Research}}
 
-## Notation
+\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
+to decompose differences in life expectancy into contributions due to
+mortality rate differences in each age. A Kitawaga (1955) decomposition
+allows us to decompose differences in a weighted mean into effects from
+differences in structure and differences in the weighted value.
 
-Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et.
-al.
-(2024), from which we here consider only symmetrical Arriaga decompositions.
-This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results.
+\textbf{Objective:} Sometimes we would like to decompose a difference
+between two populations that are each composed of like-defined
+subpopulations. Said decomposition would produce effects for each age of
+each subpopulation, as well as a marginal effect of composition
+differences. I propose a straightforward analytic method to do this.
+
+\textbf{Methods:} In short, within-subpopulation life expectancy
+differences can be handled by the Arriaga method. The case of weighting
+together the life expectancies of subpopulations by way of a composed
+radix can be handled using the well-known Kitagawa method. The elements
+of the value component of the Kitagawa method tell us how to rescale the
+Arriaga results specific to each subpopulation.
+
+\textbf{Results:} I show the proposed analytic method to be equivalent
+to a Horiuchi et al (2008) reframing of the same problem. Some
+mentionable properties: (i) There is no limit to the number of
+subpopulations, (ii) it is straightforward to incorporate cause-of-death
+information, (iii) composition is here only considered in the radix age.
+I currently have results for simulated mortality rates, but I promise to
+wrangle up an empirical application to demonstrate the method.
+
+\textbf{Conclusions:} The analytic decomposition method I propose is
+advantageous compared to a Horiuchi method for this problem purely for
+reasons of computational efficiency. This method could help further
+disentangle the effects of mortality and composition differences in
+explaining or clarifying paradoxes or secular change. I promise to think
+further about subpopulation weighting that might occur over all ages
+(using prevalence information).\}}
+
+\keywords{Decomposition, Mortality, Cause of death, Population
+Structure, Mortality Inequalities}
+
+\maketitle
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+An \citet{arriaga1984measuring} decomposition allows us to decompose
+changes in life expectancy into contributions due to mortality changes
+in different ages. The method was designed to be practical, and framed
+in terms of age lifetable columns expressed in discrete age groups. A
+well-known property of the method is that mortality changes in different
+ages need not be proportional. Derived contributions also sum exactly to
+the observed life expectancy difference. A not-well-known property of
+the method is that it is asymmetrical, in the sense that the absolute
+values of age-specific contributions depend on whether we compare
+population 1 with 2, or population 2 with 1. A further property of the
+method is that it is designed to work with homogeneous populations,
+meaning that populations 1 and 2 are each homogeneous, in the sense that
+each is expressed with only one lifetable.
+
+A \citet{kitagawa1955components} decomposition allows us to decompose
+differences in a weighted mean due to differences in weights (structure)
+and differences in the value being weighted (often rates, but in our
+case life expectancies). This widely-used decomposition method is
+well-known to be exact in that the resulting structure and value
+components sum exactly to the observed difference in weighted means. The
+individual elements (ages, or life expectancies for us) of the value
+being weighted have identifiable effects. It is not well-known that the
+individual elements of the structure component do not have identifiable
+effects. Rather, the structure effects should be summed to a a marginal
+effect due to differences in structure.
+
+Sometimes we have a situation where we would like to decompose a
+difference between two populations that are each composed of
+like-defined subpopulations. For example, life expectancy in France
+versus Spain, each with education-specific subpopulations. In such
+situations, a decomposition should tell us the contribution to the
+difference in overall life expectancy due to rate differences in each
+age in each subpopulation, and also separate an effect due to overall
+compositional change. In this paper, I would like to propose a
+straightforward analytic method to decompose in this way. In short,
+within-subpopulation life expectancy differences can be handled by the
+Arriaga method. The case of weighting together the life expectancies of
+subpopulations by way of a composed radix can be handled using the
+well-known Kitagawa method. The elements of the value component of the
+Kitagawa method tell us how to rescale the Arriaga results specific to
+each subpopulation so as to isolate the age-subpopulation-specific
+effects on the overall life expectancy difference. I will justify this
+rescaling, and show that the results of this procedure are fully
+consistent with the results of a \citet{horiuchi2008decomposition}
+reframing of the problem.
+
+\hypertarget{method}{%
+\section{Method}\label{method}}
+
+\hypertarget{notation}{%
+\subsection{Notation}\label{notation}}
+
+Arriaga decompositions can be implemented in a number of relevant ways,
+per Riffe et. al.~(2024), from which we here consider only symmetrical
+Arriaga decompositions. This implies following the original formulas
+twice, once in each direction, and averaging the sign-adjusted results.
 For this, we use the following lifetable columns:
 
-```{=tex}
 \begin{itemize}
@@
 \end{itemize}
-```
-We also use the superscript $s$ to index subpopulations comprising the total population, and the superscript $t$ to index time points, or some other way of differentiating the total population.
-The index $e^{s,t}_0$ reads as "life expectancy at birth for subpopulation $s$ at time $t$", e.g.
-$e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.
 
-Then one can write Arriaga's standard decomposition method like as per eq~\eqref{eq_arriaga1}. To keep things simple, we assume a lifetable radix of 1, meaning $\ell_0 = 1$.
+We also use the superscript \(s\) to index subpopulations comprising the
+total population, and the superscript \(t\) to index time points, or
+some other way of differentiating the total population. The index
+\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
+\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
+expectancy in 1990 of a low education group.
 
-$$
+Then one can write Arriaga's standard decomposition method like as per
+eq\textasciitilde{}\eqref{eq_arriaga1}. To keep things simple, we assume
+a lifetable radix of 1, meaning \(\ell_0 = 1\).
+
+\[
 \begin{aligned}
@@
 \end{aligned}
-$$
+\]
 
@@
 
-If your article has accompanying supplementary file/s please state so here.
+If your article has accompanying supplementary file/s please state so
+here.
 
-Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information.
-This may be requested by the editorial team/s if it is missing.
+Authors reporting data from electrophoretic gels and blots should supply
+the full unprocessed scans for key as part of their Supplementary
+information. This may be requested by the editorial team/s if it is
+missing.
 
@@
 
-Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
+Acknowledgments are not compulsory. Where included they should be brief.
+Grant or contribution numbers may be acknowledged.
 
@@
 
-# Declarations {.unnumbered}
+\hypertarget{declarations}{%
+\section*{Declarations}\label{declarations}}
+\addcontentsline{toc}{section}{Declarations}
 
-Some journals require declarations to be submitted in a standardised format.
-Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
-If yes, your manuscript must contain the following sections under the heading \`Declarations':
+Some journals require declarations to be submitted in a standardised
+format. Please check the Instructions for Authors of the journal to
+which you are submitting to see if you need to complete this section. If
+yes, your manuscript must contain the following sections under the
+heading `Declarations':
 
--   Funding
--   Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
--   Ethics approval
--   Consent to participate
--   Consent for publication
--   Availability of data and materials
--   Code availability
--   Authors' contributions
+\begin{itemize}
+\tightlist
+\item
+  Funding
+\item
+  Conflict of interest/Competing interests (check journal-specific
+  guidelines for which heading to use)
+\item
+  Ethics approval
+\item
+  Consent to participate
+\item
+  Consent for publication
+\item
+  Availability of data and materials
+\item
+  Code availability
+\item
+  Authors' contributions
+\end{itemize}
 
-\noindent If any of the sections are not relevant to your manuscript, please include the heading and write \`Not applicable' for that section.
+\noindent If any of the sections are not relevant to your manuscript,
+please include the heading and write `Not applicable' for that section.
 
-::: {.flushleft latex="true"}
+\begin{flushleft}
 Editorial Policies for:
 
-\noindent Springer journals and proceedings: <https://www.springer.com/gp/editorial-policies>
+\noindent Springer journals and proceedings:
+\url{https://www.springer.com/gp/editorial-policies}
 
-\noindent Nature Portfolio journals: <https://www.nature.com/nature-research/editorial-policies>
+\noindent Nature Portfolio journals:
+\url{https://www.nature.com/nature-research/editorial-policies}
 
-\noindent \textit{Scientific Reports}: <https://www.nature.com/srep/journal-policies/editorial-policies>
+\noindent \textit{Scientific Reports}:
+\url{https://www.nature.com/srep/journal-policies/editorial-policies}
 
-\noindent BMC journals: <https://www.biomedcentral.com/getpublished/editorial-policies>
-:::
+\noindent BMC journals:
+\url{https://www.biomedcentral.com/getpublished/editorial-policies}
 
-::: {.appendices latex="true"}
-# Section title of first appendix {#secA1}
+\end{flushleft}
 
-An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.
+\begin{appendices}
 
-For submissions to Nature Portfolio Journals please use the heading "Extended Data".
-:::
+\hypertarget{secA1}{%
+\section{Section title of first appendix}\label{secA1}}
+
+An appendix contains supplementary information that is not an essential
+part of the text itself but which may be helpful in providing a more
+comprehensive understanding of the research problem or it is information
+that is too cumbersome to be included in the body of the paper.
+
+For submissions to Nature Portfolio Journals please use the heading
+``Extended Data''.
+
+\end{appendices}
+
+\bibliography{bibliography.bib}
+
+\end{document}

### 2024-01-24 manuscript/manuscript.Rmd
updates

@@
  **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age.
-A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into effects from differences in structure and differences in the weighted value.
+A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into an effect from differences in structure and effects from differences in each element of the weighted value.
 
@@
   \newtheorem{definition}{Definition}%
-
+  \usepackage{amsmath}
 output: rticles::springer_article
@@
 al.
-(2024), from which we here consider only symmetrical Arriaga decompositions.
+(2024), from which we here consider only _symmetrical_ Arriaga decomposition.
 This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results.
@@
 $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.
+The first stage of our exercise is to decompose subgroup-specific differences in life expectancy (hence the $^s$ on everything):
 
-Then one can write Arriaga's standard decomposition method like as per eq~\eqref{eq_arriaga1}. To keep things simple, we assume a lifetable radix of 1, meaning $\ell_0 = 1$.
+\begin{align}
+{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
+ &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
+\end{align}
 
-$$
-\begin{aligned}
-{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
- &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
-\end{aligned}
-$$
+\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, ${}_n\Delta_x^s$, which can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}.
+I'll use a lifetable radix of 1, meaning $\ell_0 = 1$ to reduce the formula a bit.
+
+\begin{equation}
+\label{eq_arriaga1}
+ {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
+  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
+  \end{cases}
+\end{equation}
+
+Equation \eqref{eq_arriaga1} is the first pass of our symmetrical decomposition, where we move _forward_ in time from $t=1$ to $t=2$, so please permit . It
 
@@
 If your article has accompanying supplementary file/s please state so here.
-
-Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information.
-This may be requested by the editorial team/s if it is missing.
-
-Please refer to Journal-level guidance for any specific requirements.
 
@@
 
-::: {.flushleft latex="true"}
-Editorial Policies for:
-
-\noindent Springer journals and proceedings: <https://www.springer.com/gp/editorial-policies>
-
-\noindent Nature Portfolio journals: <https://www.nature.com/nature-research/editorial-policies>
-
-\noindent \textit{Scientific Reports}: <https://www.nature.com/srep/journal-policies/editorial-policies>
-
-\noindent BMC journals: <https://www.biomedcentral.com/getpublished/editorial-policies>
-:::
-
 ::: {.appendices latex="true"}

### 2024-01-24 manuscript/manuscript.tex
updates

@@
 \newtheorem{definition}{Definition}%
+\usepackage{amsmath}
 
@@
 mortality rate differences in each age. A Kitawaga (1955) decomposition
-allows us to decompose differences in a weighted mean into effects from
-differences in structure and differences in the weighted value.
+allows us to decompose differences in a weighted mean into an effect
+from differences in structure and effects from differences in each
+element of the weighted value.
 
@@
 Arriaga decompositions can be implemented in a number of relevant ways,
-per Riffe et. al.~(2024), from which we here consider only symmetrical
-Arriaga decompositions. This implies following the original formulas
-twice, once in each direction, and averaging the sign-adjusted results.
-For this, we use the following lifetable columns:
+per Riffe et. al.~(2024), from which we here consider only
+\emph{symmetrical} Arriaga decomposition. This implies following the
+original formulas twice, once in each direction, and averaging the
+sign-adjusted results. For this, we use the following lifetable columns:
 
@@
 \(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
-expectancy in 1990 of a low education group.
+expectancy in 1990 of a low education group. The first stage of our
+exercise is to decompose subgroup-specific differences in life
+expectancy (hence the \(^s\) on everything):
 
-Then one can write Arriaga's standard decomposition method like as per
-eq\textasciitilde{}\eqref{eq_arriaga1}. To keep things simple, we assume
-a lifetable radix of 1, meaning \(\ell_0 = 1\).
+\begin{align}
+{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
+ &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
+\end{align}
 
-\[
-\begin{aligned}
-{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
- &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
-\end{aligned}
-\]
+\noindent where \(\omega\) is the highest (open) age group, and
+\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
+expectancy being decomposed, and it is composed of age-specific
+contributions, \({}_n\Delta_x^s\), which can be calculated following
+Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll
+use a lifetable radix of 1, meaning \(\ell_0 = 1\) to reduce the formula
+a bit.
+
+\begin{equation}
+\label{eq_arriaga1}
+ {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
+  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
+  \end{cases}
+\end{equation}
+
+Equation \eqref{eq_arriaga1} is the first pass of our symmetrical
+decomposition, where we move \emph{forward} in time from \(t=1\) to
+\(t=2\), so please permit . It
 
@@
 here.
-
-Authors reporting data from electrophoretic gels and blots should supply
-the full unprocessed scans for key as part of their Supplementary
-information. This may be requested by the editorial team/s if it is
-missing.
-
-Please refer to Journal-level guidance for any specific requirements.
 
@@
 
-\begin{flushleft}
-Editorial Policies for:
-
-\noindent Springer journals and proceedings:
-\url{https://www.springer.com/gp/editorial-policies}
-
-\noindent Nature Portfolio journals:
-\url{https://www.nature.com/nature-research/editorial-policies}
-
-\noindent \textit{Scientific Reports}:
-\url{https://www.nature.com/srep/journal-policies/editorial-policies}
-
-\noindent BMC journals:
-\url{https://www.biomedcentral.com/getpublished/editorial-policies}
-
-\end{flushleft}
-
 \begin{appendices}

### 2024-01-25 manuscript/manuscript.Rmd
updates from 25 jan

@@
 
-Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et.
-al.
-(2024), from which we here consider only _symmetrical_ Arriaga decomposition.
-This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results.
-For this, we use the following lifetable columns:
+In formulas, we use the following variables and scripting, most of which are lifetable columns:
 
@@
 
+\item{$\pi^s$} the initial population fraction for subgroup $s$ at the initial age. 
 \end{itemize}
@@
 $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.
-The first stage of our exercise is to decompose subgroup-specific differences in life expectancy (hence the $^s$ on everything):
+
+## Averaging life expectancy
+
+There are two major approaches to producing average life expectancy for a total population composed of subgroups, as @vaupel2002life points out.
+One could derive the average mortality rate in each age by summing the respective deaths and exposures from subpopulations and then deriving the rate, as the Human Mortality Database @wilmoth2021methods does, then recalculate the total population lifetable from these.
+This is what Vaupel called _current rates_ approach, which ignores even controllable heterogeneity.
+The other approach derives subgroup-specific lifetables, then weights life expectancies together according to an initial mixing composition in order to arrive at the total life expectancy.
+This is Vaupel's _current conditions_ approach, at least in part.
+This second approach is what is often done in multistate models (cite), but also sometimes with standard lifetables, for example, when doing between-within decompositions of variance or other similar summary measures.
+Our decomposition method deals with this second case, specifically, we have:
+
+\begin{equation}
+\label{eq_avg}
+e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
+\end{equation}
+
+\noindent where the $S$ total subgroup fractions comprise a mixing distribution, $1 = \sum_s^S \pi^s$.
+
+## Kitagawa decomposition 
+Eq \eqref{eq_avg} treats the total life expectancy as a weighted average, which means we can decompose a change over time in $e_0$ ($\Delta = e_0^2 - e_o^1$) exactly using the formulas from @kitagawa1955components.
+Specifically, eq \eqref{eq_kit_comp} gives an overall effect of changes in composition:
+
+\begin{equation}
+\label{eq_kit_comp}
+\Delta^\text{composition} = \sum_s^S \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
+\end{equation}
+
+This result is widely known, but many are unaware that the composition effect should be summed like this, since the group-specific composition (structure) effects are not well-identified.
+Eq \eqref{eq_kit_e0} gives the subgroup-specific effects from changes in life expectancy:
+
+\begin{equation}
+\label{eq_kit_e0}
+\Delta^{e_0^s\text{~change}} = \left(e_0^{s,t2} - e_0^{s,t1}\right) \cdot \frac{\pi^{s,t1} + \pi^{s,t2}}{2}\quad \textrm{.}
+\end{equation}
+
+Eq \eqref{eq_kit_sum} says that the overall change in life expectancy is the sum of (i) a single component that captures the effect of composition change and (ii) a set of components describing the effect on overall life expectancy from each group's specific life expectancy change.
+This second effect could just as well be called the _rate_ effect, since each life expectancy is fully determined by mortality rates.
+
+\begin{equation}
+\label{eq_kit_sum}
+\Delta = \Delta^\text{composition}  + \sum_s^S \Delta^{e_0^s\text{~change}}
+\end{equation}
+
+## Symmetrical Arriaga decomposition
+The above-mentioned _rate_ effect can be conceived of as the _net_ effect on total life expectancy change of the age-specific effects within subgroups, as isolated by the many varieties of life expectancy decompositions.
+In this case, it does not matter much which method we choose to derive age-specific effects for subgroup-specific changes in life expectancy.
+The @arriaga1984measuring decomposition approach is popular in part because it is efficient, analytic, and sums exactly to the observed life expectancy difference.
+This method can be implemented in a number of relevant ways, per Riffe et.
+al.
+(2024), from which we here consider only _symmetrical_ Arriaga decomposition.
+This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results.
+We repeat this exercise for each subgroup to decompose group-specific changes in life expectancy (hence the $^s$ on everything):
 
 \begin{align}
-{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
+\label{eq_kit_e0}
+\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
  &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
@@
 
-\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, ${}_n\Delta_x^s$, which can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}.
-I'll use a lifetable radix of 1, meaning $\ell_0 = 1$ to reduce the formula a bit.
+\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, $\overrightarrow{{}_n\Delta_x^s}$ or $\overleftarrow{{}_n\Delta_x^s}$ depending on whether we decompose forward in time or backward, respectively (Riffe et.
+al., 2024).
+The forward age-specific values $\overrightarrow{{}_n\Delta_x^s}$ can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}, consistent with @arriaga1984measuring or the presentation in @preston2000demography.
+I use a lifetable radix of 1, meaning $\ell_0 = 1$, to reduce the formula slightly.
 
@@
 \label{eq_arriaga1}
- {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
+ \overrightarrow{{}_n\Delta_x^s} = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
   \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
@@
 
-Equation \eqref{eq_arriaga1} is the first pass of our symmetrical decomposition, where we move _forward_ in time from $t=1$ to $t=2$, so please permit . It
+Equation \eqref{eq_arriaga1} is the first pass of our symmetrical decomposition, and \eqref{eq_arriaga2} is the second pass, identical except all time superscripts are swapped.
+
+\begin{equation}
+\label{eq_arriaga2}
+ \overleftarrow{{}_n\Delta_x^s} = \begin{cases}\ell_x^{s,2} \cdot \left( \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}} - \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}}\right)+T_{x+n}^{s,1}\cdot\left(\frac{\ell_x^{s,2}}{\ell_x^{s,1}} - \frac{\ell_{x+n}^{s,2}}{\ell_{x+n}^{s,1}}\right)  & \text{$ \forall x < \omega$}  \\
+  \ell_\omega ^{s,2} \cdot (e_\omega^{s,1} -e_\omega^{s,2} ) & \text{$\forall x = \omega $}
+  \end{cases}
+\end{equation}
+
+Importantly,
+
+\begin{align}
+\label{eq_sum2}
+ -\Delta^s &= \sum_x^\omega \overleftarrow{{}_n\Delta_x^s} \\
+ &= e_0^1 - e_0^2 \nonumber
+\end{align}
+
+Then our symmetrical estimate of ${}_n\Delta_x^s$ is just the average of these:
+
+\begin{equation}
+\label{eq_arriaga3}
+{}_n\Delta_x^s = \frac{\left(\overrightarrow{{}_n\Delta_x^s} -  \overleftarrow{{}_n\Delta_x^s}\right)}{2}
+\end{equation}
+
+Repeat the above to derive ${}_n\Delta_x^s$ for each within-subpopulation comparison.
+
+## Rescale Arriaga results
+
+The quantity ${}_n\Delta_x^s$ given in eq \eqref{eq_arriaga3} is on the scale of group $s$ specific life expectancy changes, but to know its net impact on overall life expectancy change we rescale per eq \eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}
+
+\begin{equation}
+\label{eq_rescale}
+{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
+\end{equation}
 

### 2024-01-25 manuscript/manuscript.tex
updates from 25 jan

@@
 
-Arriaga decompositions can be implemented in a number of relevant ways,
-per Riffe et. al.~(2024), from which we here consider only
-\emph{symmetrical} Arriaga decomposition. This implies following the
-original formulas twice, once in each direction, and averaging the
-sign-adjusted results. For this, we use the following lifetable columns:
+In formulas, we use the following variables and scripting, most of which
+are lifetable columns:
 
@@
 \(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
-expectancy in 1990 of a low education group. The first stage of our
-exercise is to decompose subgroup-specific differences in life
-expectancy (hence the \(^s\) on everything):
+expectancy in 1990 of a low education group.
+
+\hypertarget{averaging-life-expectancy}{%
+\subsection{Averaging life expectancy}\label{averaging-life-expectancy}}
+
+There are two major approaches to producing average life expectancy for
+a total population composed of subgroups, as \citet{vaupel2002life}
+points out. One could derive the average mortality rate in each age by
+summing the respective deaths and exposures from subpopulations and then
+deriving the rate, as the Human Mortality Database
+\citet{wilmoth2021methods} does, then recalculate the total population
+lifetable from these. This is what Vaupel called \emph{current rates}
+approach, which ignores even controllable heterogeneity. The other
+approach derives subgroup-specific lifetables, then weights life
+expectancies together according to an initial mixing composition in
+order to arrive at the total life expectancy. This is Vaupel's
+\emph{current conditions} approach, at least in part. This second
+approach is what is often done in multistate models (cite), but also
+sometimes with standard lifetables, for example, when doing
+between-within decompositions of variance or other similar summary
+measures. Our decomposition method deals with this second case,
+specifically, we have:
+
+\begin{equation}
+\label{eq_avg}
+e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
+\end{equation}
+
+\noindent where the \(S\) total subgroup fractions comprise a mixing
+distribution, \(1 = \sum_s^S \pi^s\).
+
+\hypertarget{kitagawa-decomposition}{%
+\subsection{Kitagawa decomposition}\label{kitagawa-decomposition}}
+
+Eq \eqref{eq_avg} treats the total life expectancy as a weighted
+average, which means we can decompose a change over time in \(e_0\)
+(\(\Delta = e_0^2 - e_o^1\)) exactly using the formulas from
+\citet{kitagawa1955components}. Specifically, eq \eqref{eq_kit_comp}
+gives an overall effect of changes in composition:
+
+\begin{equation}
+\label{eq_kit_comp}
+\Delta^\text{composition} = \sum_s^S \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
+\end{equation}
+
+This result is widely known, but many are unaware that the composition
+effect should be summed like this, since the group-specific composition
+(structure) effects are not well-identified. Eq \eqref{eq_kit_e0} gives
+the subgroup-specific effects from changes in life expectancy:
+
+\begin{equation}
+\label{eq_kit_e0}
+\Delta^{e_0^s\text{~change}} = \left(e_0^{s,t2} - e_0^{s,t1}\right) \cdot \frac{\pi^{s,t1} + \pi^{s,t2}}{2}\quad \textrm{.}
+\end{equation}
+
+Eq \eqref{eq_kit_sum} says that the overall change in life expectancy is
+the sum of (i) a single component that captures the effect of
+composition change and (ii) a set of components describing the effect on
+overall life expectancy from each group's specific life expectancy
+change. This second effect could just as well be called the \emph{rate}
+effect, since each life expectancy is fully determined by mortality
+rates.
+
+\begin{equation}
+\label{eq_kit_sum}
+\Delta = \Delta^\text{composition}  + \sum_s^S \Delta^{e_0^s\text{~change}}
+\end{equation}
+
+\hypertarget{symmetrical-arriaga-decomposition}{%
+\subsection{Symmetrical Arriaga
+decomposition}\label{symmetrical-arriaga-decomposition}}
+
+The above-mentioned \emph{rate} effect can be conceived of as the
+\emph{net} effect on total life expectancy change of the age-specific
+effects within subgroups, as isolated by the many varieties of life
+expectancy decompositions. In this case, it does not matter much which
+method we choose to derive age-specific effects for subgroup-specific
+changes in life expectancy. The \citet{arriaga1984measuring}
+decomposition approach is popular in part because it is efficient,
+analytic, and sums exactly to the observed life expectancy difference.
+This method can be implemented in a number of relevant ways, per Riffe
+et. al.~(2024), from which we here consider only \emph{symmetrical}
+Arriaga decomposition. This implies following the original formulas
+twice, once in each direction, and averaging the sign-adjusted results.
+We repeat this exercise for each subgroup to decompose group-specific
+changes in life expectancy (hence the \(^s\) on everything):
 
 \begin{align}
-{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
+\label{eq_kit_e0}
+\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
  &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
@@
 expectancy being decomposed, and it is composed of age-specific
-contributions, \({}_n\Delta_x^s\), which can be calculated following
-Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll
-use a lifetable radix of 1, meaning \(\ell_0 = 1\) to reduce the formula
-a bit.
+contributions, \(\overrightarrow{{}_n\Delta_x^s}\) or
+\(\overleftarrow{{}_n\Delta_x^s}\) depending on whether we decompose
+forward in time or backward, respectively (Riffe et. al., 2024). The
+forward age-specific values \(\overrightarrow{{}_n\Delta_x^s}\) can be
+calculated following Arriaga's decomposition method following eq
+\eqref{eq_arriaga1}, consistent with \citet{arriaga1984measuring} or the
+presentation in \citet{preston2000demography}. I use a lifetable radix
+of 1, meaning \(\ell_0 = 1\), to reduce the formula slightly.
 
@@
 Equation \eqref{eq_arriaga1} is the first pass of our symmetrical
-decomposition, where we move \emph{forward} in time from \(t=1\) to
-\(t=2\), so please permit . It
+decomposition, and \eqref{eq_arriaga2} is the second pass, identical
+except all time superscripts are swapped.
+
+\begin{equation}
+\label{eq_arriaga2}
+ \overleftarrow{{}_n\Delta_x^s} = \begin{cases}\ell_x^{s,2} \cdot \left( \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}} - \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}}\right)+T_{x+n}^{s,1}\cdot\left(\frac{\ell_x^{s,2}}{\ell_x^{s,1}} - \frac{\ell_{x+n}^{s,2}}{\ell_{x+n}^{s,1}}\right)  & \text{$ \forall x < \omega$}  \\
+  \ell_\omega ^{s,2} \cdot (e_\omega^{s,1} -e_\omega^{s,2} ) & \text{$\forall x = \omega $}
+  \end{cases}
+\end{equation}
+
+Importantly,
+
+\begin{align}
+\label{eq_sum2}
+ -\Delta^s &= \sum_x^\omega \overleftarrow{{}_n\Delta_x^s} \\
+ &= e_0^1 - e_0^2 \nonumber
+\end{align}
+
+Then our symmetrical estimate of \({}_n\Delta_x^s\) is just the average
+of these:
+
+\begin{equation}
+\label{eq_arriaga3}
+{}_n\Delta_x^s = \frac{\left(\overrightarrow{{}_n\Delta_x^s} -  \overleftarrow{{}_n\Delta_x^s}\right)}{2}
+\end{equation}
+
+Repeat the above to derive \({}_n\Delta_x^s\) for each
+within-subpopulation comparison.
+
+\hypertarget{rescale-arriaga-results}{%
+\subsection{Rescale Arriaga results}\label{rescale-arriaga-results}}
+
+The quantity \({}_n\Delta_x^s\) given in eq \eqref{eq_arriaga3} is on
+the scale of group \(s\) specific life expectancy changes, but to know
+its net impact on overall life expectancy change we rescale per eq
+\eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}
+
+\begin{equation}
+\label{eq_rescale}
+{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
+\end{equation}
 

### 2024-01-25 manuscript/manuscript.Rmd
late updates; time to find an empirical example

@@
 
-# Introduction
+# Background
 
@@
 
-Sometimes we have a situation where we would like to decompose a difference between two populations that are each composed of like-defined subpopulations.
-For example, life expectancy in France versus Spain, each with education-specific subpopulations.
-In such situations, a decomposition should tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change.
+# Introduction
+A demographer might like to decompose a difference in life expectancy between two populations (or one population at two times) that are each composed of like-defined subpopulations.
+For example, life expectancy in France versus Spain, each with education-specific subpopulations, or a national life expectancy calculated as the weighted average of its regions at two points in time.
+In such situations, a decomposition would ideally tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change.
 In this paper, I would like to propose a straightforward analytic method to decompose in this way.
@@
 \label{eq_avg}
-e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
+e_0 = \sum_s \pi^s \cdot e_0^s \quad \textrm{,}
 \end{equation}
 
-\noindent where the $S$ total subgroup fractions comprise a mixing distribution, $1 = \sum_s^S \pi^s$.
+\noindent where the $S$ total subgroup fractions comprise a mixing distribution, $1 = \sum_s \pi^s$.
 
@@
 \label{eq_kit_comp}
-\Delta^\text{composition} = \sum_s^S \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
+\Delta^\text{composition} = \sum_s \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
 \end{equation}
@@
 \label{eq_kit_sum}
-\Delta = \Delta^\text{composition}  + \sum_s^S \Delta^{e_0^s\text{~change}}
+\Delta = \Delta^\text{composition}  + \sum_s \Delta^{e_0^s\text{~change}}
 \end{equation}
@@
 \Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
- &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
+ &= \sum _x {}_n\Delta_x^s \quad \textrm{,}\nonumber 
 \end{align}
 
-\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, $\overrightarrow{{}_n\Delta_x^s}$ or $\overleftarrow{{}_n\Delta_x^s}$ depending on whether we decompose forward in time or backward, respectively (Riffe et.
+\noindent where $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, $\overrightarrow{{}_n\Delta_x^s}$ or $\overleftarrow{{}_n\Delta_x^s}$ depending on whether we decompose forward in time or backward, respectively (Riffe et.
 al., 2024).
@@
 \label{eq_sum2}
- -\Delta^s &= \sum_x^\omega \overleftarrow{{}_n\Delta_x^s} \\
+ -\Delta^s &= \sum_x \overleftarrow{{}_n\Delta_x^s} \\
  &= e_0^1 - e_0^2 \nonumber
@@
 
-The quantity ${}_n\Delta_x^s$ given in eq \eqref{eq_arriaga3} is on the scale of group $s$ specific life expectancy changes, but to know its net impact on overall life expectancy change we rescale per eq \eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}
+The age-specific quantities ${}_n\Delta_x^s$ derived from eq \eqref{eq_arriaga3} are on the scale of subgroup-specific life expectancy changes.
+To know the net impact of ${}_n\Delta_x^s$ on overall life expectancy change we rescale per eq \eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}
 
@@
 \label{eq_rescale}
-{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
+{}_n\Delta_x^{s,\text{net}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s} \quad \mathrm{.}
+\end{equation}
+
+With eqs \eqref{eq_kit_comp} and \eqref{eq_rescale}, we have all elements of the present decomposition:
+
+\begin{equation}
+\label{eq_full}
+\Delta = \Delta^\text{composition} + \sum_s \sum_x {}_n\Delta_x^{s,\text{net}}
 \end{equation}
@@
 
-Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
+Thanks to Michael Lachanski for posing to me the question that led to the solution I present here, to Jonas Schoeley for first bringing the composition problem with the Kitagawa method to my attention, and to the MPIDR Laboratory of Population Health for posing assorted questions in September 2023 that got me thinking in these lines.
 
@@
 Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
-If yes, your manuscript must contain the following sections under the heading \`Declarations':
+If yes, your manuscript must contain the following sections under the heading
 
--   Funding
--   Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
--   Ethics approval
--   Consent to participate
--   Consent for publication
--   Availability of data and materials
--   Code availability
--   Authors' contributions
+## Funding
+This work was financed in part by grant nr SR22-00502 from _la Caixa Foundation_ and grant nr PID2022-142762OA-I00 from the Spanish _Ministerio de Ciencia e Innovación_.
 
-\noindent If any of the sections are not relevant to your manuscript, please include the heading and write \`Not applicable' for that section.
+## Conflict of interest
+The author declares no conflict of interest
+
+## Availability of data and materials
+All data are ___
+
+## Code availability
+All code used here is available in an open `GitHub` repository \url{https://github.com/timriffe/arriagakitagawa}
 

### 2024-01-25 manuscript/manuscript.tex
late updates; time to find an empirical example

@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
@@
 
-Sometimes we have a situation where we would like to decompose a
-difference between two populations that are each composed of
-like-defined subpopulations. For example, life expectancy in France
-versus Spain, each with education-specific subpopulations. In such
-situations, a decomposition should tell us the contribution to the
-difference in overall life expectancy due to rate differences in each
-age in each subpopulation, and also separate an effect due to overall
-compositional change. In this paper, I would like to propose a
-straightforward analytic method to decompose in this way. In short,
-within-subpopulation life expectancy differences can be handled by the
-Arriaga method. The case of weighting together the life expectancies of
-subpopulations by way of a composed radix can be handled using the
-well-known Kitagawa method. The elements of the value component of the
-Kitagawa method tell us how to rescale the Arriaga results specific to
-each subpopulation so as to isolate the age-subpopulation-specific
-effects on the overall life expectancy difference. I will justify this
-rescaling, and show that the results of this procedure are fully
-consistent with the results of a \citet{horiuchi2008decomposition}
-reframing of the problem.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+A demographer might like to decompose a difference in life expectancy
+between two populations (or one population at two times) that are each
+composed of like-defined subpopulations. For example, life expectancy in
+France versus Spain, each with education-specific subpopulations, or a
+national life expectancy calculated as the weighted average of its
+regions at two points in time. In such situations, a decomposition would
+ideally tell us the contribution to the difference in overall life
+expectancy due to rate differences in each age in each subpopulation,
+and also separate an effect due to overall compositional change. In this
+paper, I would like to propose a straightforward analytic method to
+decompose in this way. In short, within-subpopulation life expectancy
+differences can be handled by the Arriaga method. The case of weighting
+together the life expectancies of subpopulations by way of a composed
+radix can be handled using the well-known Kitagawa method. The elements
+of the value component of the Kitagawa method tell us how to rescale the
+Arriaga results specific to each subpopulation so as to isolate the
+age-subpopulation-specific effects on the overall life expectancy
+difference. I will justify this rescaling, and show that the results of
+this procedure are fully consistent with the results of a
+\citet{horiuchi2008decomposition} reframing of the problem.
 
@@
 \label{eq_avg}
-e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
+e_0 = \sum_s \pi^s \cdot e_0^s \quad \textrm{,}
 \end{equation}
@@
 \noindent where the \(S\) total subgroup fractions comprise a mixing
-distribution, \(1 = \sum_s^S \pi^s\).
+distribution, \(1 = \sum_s \pi^s\).
 
@@
 \Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
- &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
+ &= \sum _x {}_n\Delta_x^s \quad \textrm{,}\nonumber 
 \end{align}
 
-\noindent where \(\omega\) is the highest (open) age group, and
-\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
-expectancy being decomposed, and it is composed of age-specific
+\noindent where \(\Delta^s\) is the subgroup-specific (\(s\)) difference
+in life expectancy being decomposed, and it is composed of age-specific
 contributions, \(\overrightarrow{{}_n\Delta_x^s}\) or
@@
 
-The quantity \({}_n\Delta_x^s\) given in eq \eqref{eq_arriaga3} is on
-the scale of group \(s\) specific life expectancy changes, but to know
-its net impact on overall life expectancy change we rescale per eq
-\eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}
+The age-specific quantities \({}_n\Delta_x^s\) derived from eq
+\eqref{eq_arriaga3} are on the scale of subgroup-specific life
+expectancy changes. To know the net impact of \({}_n\Delta_x^s\) on
+overall life expectancy change we rescale per eq \eqref{eq_rescale} to
+match the effect sizes from eq \eqref{eq_kit_e0}
 
@@
 \label{eq_rescale}
-{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
+{}_n\Delta_x^{s,\text{net}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s} \quad \mathrm{.}
+\end{equation}
+
+With eqs \eqref{eq_kit_comp} and \eqref{eq_rescale}, we have all
+elements of the present decomposition:
+
+\begin{equation}
+\label{eq_full}
+\Delta = \Delta^\text{composition} + \sum_s \sum_x {}_n\Delta_x^{s,\text{net}}
 \end{equation}
@@
 
-Acknowledgments are not compulsory. Where included they should be brief.
-Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
+Thanks to Michael Lachanski for posing to me the question that led to
+the solution I present here, to Jonas Schoeley for first bringing the
+composition problem with the Kitagawa method to my attention, and to the
+MPIDR Laboratory of Population Health for posing assorted questions in
+September 2023 that got me thinking in these lines.
 
@@
 yes, your manuscript must contain the following sections under the
-heading `Declarations':
+heading
 
-\begin{itemize}
-\tightlist
-\item
-  Funding
-\item
-  Conflict of interest/Competing interests (check journal-specific
-  guidelines for which heading to use)
-\item
-  Ethics approval
-\item
-  Consent to participate
-\item
-  Consent for publication
-\item
-  Availability of data and materials
-\item
-  Code availability
-\item
-  Authors' contributions
-\end{itemize}
+\hypertarget{funding}{%
+\subsection{Funding}\label{funding}}
 
-\noindent If any of the sections are not relevant to your manuscript,
-please include the heading and write `Not applicable' for that section.
+This work was financed in part by grant nr SR22-00502 from \emph{la
+Caixa Foundation} and grant nr PID2022-142762OA-I00 from the Spanish
+\emph{Ministerio de Ciencia e Innovación}.
+
+\hypertarget{conflict-of-interest}{%
+\subsection{Conflict of interest}\label{conflict-of-interest}}
+
+The author declares no conflict of interest
+
+\hypertarget{availability-of-data-and-materials}{%
+\subsection{Availability of data and
+materials}\label{availability-of-data-and-materials}}
+
+All data are \_\_\_
+
+\hypertarget{code-availability}{%
+\subsection{Code availability}\label{code-availability}}
+
+All code used here is available in an open \texttt{GitHub} repository
+\url{https://github.com/timriffe/arriagakitagawa}
 

### 2024-09-12 R/01_smoothing_and_ungroupping.R
Final commit. Comments attended figures slightly modified. Post covid figures moved to separate script

@@ -1,7 +1,7 @@
 # initial data
 source("R/00_initial_data_preparation.R")
 # ----------------------------------------------------------------------- #
-# I will use this data further for predict() with GAM model results
+# Data to further predict() with GAM model results
 new_data <- expand_grid(
   age  = seq(35, 100, 1),
   year  = c("2016-2019", "2020-2021"),

### 2024-09-12 R/02_LT_and_average_quantities.R

@@ -9,7 +9,7 @@ Lt <- mxc_single |>
   group_by(sex, educ, cause, year) |>  
   summarise(ex = mx_to_e0(mx, age = age), .groups = "drop")
 
-# we compare this later with our weighted-average e35 values just to see.
+# we compare this later with our weighted-average e35 values
 e35_total_compare <- Lt |>
   filter(educ == "Total", sex != "Total") |> 
   select(sex, year, ex)

### 2024-09-12 R/03_decomposition_results.R

@@ -55,8 +55,8 @@ struct_kit <- data_5_prepped |>
 kit <- left_join(struct_kit, 
                  e35_kit, 
                  by = join_by(educ, year)) |> 
-  mutate(e35_component = st_mean * e35_diff, #5 ############
-         st_component  = e35_avg * st_diff) #3
+  mutate(e35_component = st_mean * e35_diff,
+         st_component  = e35_avg * st_diff)
 
 decomp_total <- kit |> 
   select(educ, year, e35_component) |> 

### 2024-09-12 R/04_gaps.R

@@ -10,7 +10,6 @@ gaps <- kit |>
   mutate(gap = e35_Females - e35_Males)
 
 tot_gps <- gaps |> 
-  # filter(year == "2016-2019") |> 
   dplyr::select(year, Females = e35_Females, Males = e35_Males) |> 
   pivot_longer(-year,
                names_to = "sex",
@@ -20,15 +19,13 @@ tot_gps <- gaps |>
 ed_gps <- e35_kit |> 
   select(educ, year, Females = e35_Females, Males = e35_Males)  |>      
   pivot_longer(Females:Males, names_to = "sex", values_to = "e35")
-# |> 
-#   filter(year == "2016-2019")
 
-# table
+# table column
 decomp_total |>
   group_by(year, educ) |>
-  summarize(cc_mort = sum(result_rescaled)) %>%
-  filter(year == "2016-2019") %>%
-  pull(cc_mort) %>%
+  summarize(cc_mort = sum(result_rescaled)) |> 
+  filter(year == "2016-2019") |> 
+  pull(cc_mort) |> 
   sum()
 
 mort_gaps <- decomp_total |> 
@@ -45,7 +42,6 @@ dec_gaps <- full_join(mort_gaps, st_gaps, by = join_by(year)) |>
 st_bind <- kit |> 
   ungroup() |> 
   group_by(year) |> 
-  # filter(year == "2016-2019") |>
   summarize(margin = sum(st_component)) |> 
   mutate(educ = "Educ. Composition", .before = 1)
 
@@ -56,12 +52,10 @@ non_stationary_gap <-
 
 education <- e35_kit |> 
   select(educ, year, e35_diff)  |>
-  # filter(year == "2016-2019") |> 
   mutate(type = "By education") |> 
   rename(gap = e35_diff)
 
 tot_gps <- gaps |> 
-  # filter(year == "2016-2019") |> 
   dplyr::select(year, gap) |> 
   mutate(educ = "Total") |> 
   mutate(type = "Stationary")
@@ -69,7 +63,6 @@ tot_gps <- gaps |>
 orig_gap <- e35_total_compare |> 
   pivot_wider(names_from = sex, values_from = ex, names_prefix = "e35_") |> 
   mutate(gap = e35_Females - e35_Males) |>  
-  # filter(year == "2016-2019") |> 
   dplyr::select(year, gap) |> 
   mutate(type = "Non-Stationary") |> 
   mutate(educ = "Total")

### 2024-09-12 R/decompose.R
.

@@ -1,9 +1,0 @@
-
-library(coddecomp)
-library(tidyverse)
-
-mxc <- read_csv("data/mxc.csv") |> 
-  filter(year == 2019)
-
-head(mxc)
-

### 2024-09-12 manuscript/manuscript.tex

--- 2024-01-24 manuscript/manuscript.tex
@@
-%Version 2.1 April 2023
-% See section 11 of the User Manual for version history
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[sn-apa,pdflatex]{sn-jnl}
+\documentclass[12pt, sn-apa,pdflatex,letterpaper]{sn-jnl}
 
@@
 
+\usepackage[T1]{fontenc}
+\usepackage{tgbonum}
 \usepackage{graphicx}%
@@
 \usepackage{listings}%
+\usepackage{natbib}
+\usepackage[pass]{geometry}
+\usepackage{makecell}
+
+\usepackage{lineno}
+\linenumbers
+
 %%%%
-
-%%%%%=============================================================================%%%%
-%%%%  Remarks: This template is provided to aid authors with the preparation
-%%%%  of original research articles intended for submission to journals published
-%%%%  by Springer Nature. The guidance has been prepared in partnership with
-%%%%  production teams to conform to Springer Nature technical requirements.
-%%%%  Editorial and presentation requirements differ among journal portfolios and
-%%%%  research disciplines. You may find sections in this template are irrelevant
-%%%%  to your work and are empowered to omit any such section if allowed by the
-%%%%  journal you intend to submit to. The submission guidelines and policies
-%%%%  of the journal take precedence. A detailed User Manual is available in the
-%%%%  template package for technical guidance.
-%%%%%=============================================================================%%%%
-
-%% Per the spinger doc, new theorem styles can be included using built in style, 
-%% but it seems the don't work so commented below
-%\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-
-%% \theoremstyle{thmstyletwo}%
-\theoremstyle{remark}
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-
-%% \theoremstyle{thmstylethree}%
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}%
-\usepackage{amsmath}
-
 \raggedbottom
-
-% tightlist command for lists without linebreak
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\renewcommand{\baselinestretch}{1.5}
 
@@
 
-\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. life expectancy
-decompositions including population subgroups}
-
-%%=============================================================%%
-%% Prefix	-> \pfx{Dr}
-%% GivenName	-> \fnm{Joergen W.}
-%% Particle	-> \spfx{van der} -> surname prefix
-%% FamilyName	-> \sur{Ploeg}
-%% Suffix	-> \sfx{IV}
-%% NatureName	-> \tanm{Poet Laureate} -> Title after name
-%% Degrees	-> \dgr{MSc, PhD}
-%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
-%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
-%%=============================================================%%
+\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. Life expectancy decomposition and population subgroups}
 
@@
 
-  \affil*[1]{\orgdiv{Department of Sociology and Social
+\author*[1]{ \fnm{Rustam} \sur{Tursun-zade} }
+
+\author*[4]{\pfx{Dr.}  \fnm{Sergi} \sur{Trias Llimós} }
+
+  \affil*[1]{\orgdiv{OPIK, Department of Sociology and Social
 Work}, \orgname{University of the Basque Country
@@
 Planck Institute for Demographic Research}}
+\affil[4]{ \orgname{Centre for Demographic Studies}}
 
-\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
-to decompose differences in life expectancy into contributions due to
-mortality rate differences in each age. A Kitawaga (1955) decomposition
-allows us to decompose differences in a weighted mean into an effect
-from differences in structure and effects from differences in each
-element of the weighted value.
+\doublespacing
 
-\textbf{Objective:} Sometimes we would like to decompose a difference
-between two populations that are each composed of like-defined
-subpopulations. Said decomposition would produce effects for each age of
-each subpopulation, as well as a marginal effect of composition
-differences. I propose a straightforward analytic method to do this.
+\abstract{\textbf{Background}: An Arriaga (1984) decomposition partitions differences in life expectancy into contributions from mortality rate differences in each age.
+A Kitagawa (1955) decomposition partitions a difference between two weighted means into effects from differences in structure and from differences in each element of the weighted value.
 
-\textbf{Methods:} In short, within-subpopulation life expectancy
-differences can be handled by the Arriaga method. The case of weighting
-together the life expectancies of subpopulations by way of a composed
-radix can be handled using the well-known Kitagawa method. The elements
-of the value component of the Kitagawa method tell us how to rescale the
-Arriaga results specific to each subpopulation.
+\textbf{Objective:} To propose an analytic method to decompose the difference in life expectancy between two populations, each consisting of like-defined subpopulations.
+Said decomposition would yield effects for each age within each subpopulation and a marginal effect that accounts for composition differences.
+We illustrate it by examining the sex gap in life expectancy in Spain using education-specific mortality data.
 
-\textbf{Results:} I show the proposed analytic method to be equivalent
-to a Horiuchi et al (2008) reframing of the same problem. Some
-mentionable properties: (i) There is no limit to the number of
-subpopulations, (ii) it is straightforward to incorporate cause-of-death
-information, (iii) composition is here only considered in the radix age.
-I currently have results for simulated mortality rates, but I promise to
-wrangle up an empirical application to demonstrate the method.
+\textbf{Methods:} Life expectancy differences between like-defined subpopulations can be decomposed using the Arriaga method, or a variant thereof.
+The results of subgroup-specific decompositions can be weighted together using the value component from a Kitagawa decomposition of total life expectancy, given that subpopulations are blended in the radix age.
+The composition component of the same Kitagawa decomposition gives the effect of subgroup composition on total life expectancy differences.
 
-\textbf{Conclusions:} The analytic decomposition method I propose is
-advantageous compared to a Horiuchi method for this problem purely for
-reasons of computational efficiency. This method could help further
-disentangle the effects of mortality and composition differences in
-explaining or clarifying paradoxes or secular change. I promise to think
-further about subpopulation weighting that might occur over all ages
-(using prevalence information).\}}
+\textbf{Results:} Notable properties of the method include: (i) it accommodates any number of subpopulations, (ii) it easily incorporates cause-of-death information, and (iii) composition is considered in the radix age.
+We apply the method to Spanish cause- and education-specific data.
+We demonstrate that the proposed analytic method is equivalent to reframing the same problem using the generalized linear integral decomposition method.
 
-\keywords{Decomposition, Mortality, Cause of death, Population
-Structure, Mortality Inequalities}
+\textbf{Conclusions:} This method can further disentangle the effects of mortality and composition differences, helping to explain or clarify paradoxes and secular change.
+Our analytic decomposition method is more computationally efficient than the linear integral or stepwise replacement decomposition methods.
+We give both R code and spreadsheet implementations of the method.}
+
+\keywords{Decomposition, Mortality, Cause of death, Population Structure, Mortality Inequalities}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section*{Introduction}\label{intro}
 
-An \citet{arriaga1984measuring} decomposition allows us to decompose
-changes in life expectancy into contributions due to mortality changes
-in different ages. The method was designed to be practical, and framed
-in terms of age lifetable columns expressed in discrete age groups. A
-well-known property of the method is that mortality changes in different
-ages need not be proportional. Derived contributions also sum exactly to
-the observed life expectancy difference. A not-well-known property of
-the method is that it is asymmetrical, in the sense that the absolute
-values of age-specific contributions depend on whether we compare
-population 1 with 2, or population 2 with 1. A further property of the
-method is that it is designed to work with homogeneous populations,
-meaning that populations 1 and 2 are each homogeneous, in the sense that
-each is expressed with only one lifetable.
+Population-level differences in life expectancy have been widely analyzed using decomposition methods to understand the impacts of mortality differences in age groups \citep[e.g.,][]{arriaga1984measuring, pollard1988decomposition},  or \citet{kitagawa1955components} to understand better the impact of compositional differences in social strata \citep{kitagawa1977mortality}.
+It has long been known among demographers that improvements in social welfare have the potential to bring large scale improvements in mortality conditions even net of medical improvements \citep{kitagawa1968education}.
+However, very few demographers have ventured to account for both age-specific effects and compositional effects at the same time \citep{torres2019contribution, hendi2021immigration, su2023educational, su2024subnational},.
+Disentangling the compositional effect within a conventional age group decomposition would improve our understanding of life expectancy differences between populations.
 
-A \citet{kitagawa1955components} decomposition allows us to decompose
-differences in a weighted mean due to differences in weights (structure)
-and differences in the value being weighted (often rates, but in our
-case life expectancies). This widely-used decomposition method is
-well-known to be exact in that the resulting structure and value
-components sum exactly to the observed difference in weighted means. The
-individual elements (ages, or life expectancies for us) of the value
-being weighted have identifiable effects. It is not well-known that the
-individual elements of the structure component do not have identifiable
-effects. Rather, the structure effects should be summed to a a marginal
-effect due to differences in structure.
+The life expectancy decomposition method proposed by \citet{arriaga1984measuring} separates a change in life expectancy into contributions due to mortality changes at different ages.
+This technique is designed to be practical in that it is framed in terms of lifetable columns expressed in discrete age groups.
+Two well-known properties of the method are that mortality changes in different ages do not need to be proportional and that the derived contributions sum exactly to the observed difference in life expectancy.
+A documented but lesser-known property of this method is its asymmetry: the absolute values of age-specific contributions depend on whether we compare population one with two, or vice versa.
+This is why the method is described in terms of changes rather than group differences; the direction of time is clear, so we always decompose an earlier period against a later period.
+Importantly, the method is designed to work with homogeneous populations, meaning a single lifetable represents each population.
+Let's call this method and the variant of it that we later apply the ``Arriaga method''.
+The Arriaga method has different extensions available to separate the effects of different causes of death within age groups.
+We seek to extend this method to account for lifetables that are composed of subpopulations with different mortality.
 
-Sometimes we have a situation where we would like to decompose a
-difference between two populations that are each composed of
-like-defined subpopulations. For example, life expectancy in France
-versus Spain, each with education-specific subpopulations. In such
-situations, a decomposition should tell us the contribution to the
-difference in overall life expectancy due to rate differences in each
-age in each subpopulation, and also separate an effect due to overall
-compositional change. In this paper, I would like to propose a
-straightforward analytic method to decompose in this way. In short,
-within-subpopulation life expectancy differences can be handled by the
-Arriaga method. The case of weighting together the life expectancies of
-subpopulations by way of a composed radix can be handled using the
-well-known Kitagawa method. The elements of the value component of the
-Kitagawa method tell us how to rescale the Arriaga results specific to
-each subpopulation so as to isolate the age-subpopulation-specific
-effects on the overall life expectancy difference. I will justify this
-rescaling, and show that the results of this procedure are fully
-consistent with the results of a \citet{horiuchi2008decomposition}
-reframing of the problem.
+The decomposition method proposed by \citet{kitagawa1955components} separates differences in a weighted mean into contributions from differences in weights (structure) and the value being weighted.
+Often, the weighted values are rates, but in our case, the weighted values will be the life expectancies of subgroups.
+This method is well-known to partition differences owing to ``structure'' and ``value'' components, which sum exactly to the observed difference in weighted means.
+The individual elements of the value being weighted (in our case, life expectancies) have identifiable effects.
+It is not well-known that the individual elements of the structure component do not have identifiable effects.
+Instead, the structure effects should be summed up as a total marginal effect due to composition differences.
+We refer to this method as the ``Kitagawa'' method.
 
-\hypertarget{method}{%
-\section{Method}\label{method}}
+We propose to treat the problem of life expectancy decomposition by age, where a total life expectancy is calculated as the weighted mean of the expectancies of subpopulations.
+This is only practical when each population is composed of like-defined subpopulations.
+For example, we may decompose the life expectancy difference between two countries considering differences in education structure, since life expectancy varies by education level \citep{mackenbach2019determinants, trias2023age}.
+Or, one might wish to calculate a national life expectancy as the weighted average of its regions at two points in time since regional weightings and inequalities may have changed over time.
+In such scenarios, a decomposition should yield the contribution to the overall life expectancy difference stemming from rate differences in each age (and potentially causes of death) within each subpopulation while giving the overall effect of compositional differences.
+We propose a straightforward analytic method to perform such decompositions.
 
-\hypertarget{notation}{%
-\subsection{Notation}\label{notation}}
+In short, differences in life expectancy between subpopulations can be decomposed using the Arriaga method, one of its variants, or by any other life expectancy decomposition method that is acceptably precise with discrete data \citep{Ponnapalli2005,horiuchi2008decomposition,andreev2002algorithm}.
+In a second step, the Kitagawa method is then used to rescale group-specific life expectancy decompositions, yielding an extra component telling us the importance of compositional differences.
+We justify this second-pass rescaling and demonstrate that the results of this procedure align with those of a linear integral reframing of the problem, per \citet{horiuchi2008decomposition}.
+We then apply these methods to decompose the sex gap between Spanish men and women in remaining life expectancy at age 35, considering differences in composition by education attainment around the same age and considering cause-specific mortality rates.
 
-Arriaga decompositions can be implemented in a number of relevant ways,
-per Riffe et. al.~(2024), from which we here consider only
-\emph{symmetrical} Arriaga decomposition. This implies following the
-original formulas twice, once in each direction, and averaging the
-sign-adjusted results. For this, we use the following lifetable columns:
+\section*{Method}\label{method}
+
+\subsection*{Notation}\label{notation}
+
+We use the following variables and scripting, most of which are lifetable columns:
 
 \begin{itemize}
+\item{${}_nm_a$} mortality rate in the age interval $[a,a+n)$.
 
-\item{$\ell(a)$} lifetable survivorship at exact age $a$.
+\item{$\ell_a$} lifetable survivorship at exact age $a$.
 
-\item{${}_nL_a$} lifetable person years lived in the interval $[a,a+n)$.
+\item{${}_nL_a$} lifetable person-years lived in the interval $[a,a+n)$.
 
-\item{$T_a$} total lifetable person years lived beyond age $a$.
+\item{$T_a$} total lifetable person-years lived beyond age $a$.
 
@@
 
+\item{$\pi_a^s$} the population fraction for subgroup $s$ in age $a$.
 \end{itemize}
 
-We also use the superscript \(s\) to index subpopulations comprising the
-total population, and the superscript \(t\) to index time points, or
-some other way of differentiating the total population. The index
-\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
-\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
-expectancy in 1990 of a low education group. The first stage of our
-exercise is to decompose subgroup-specific differences in life
-expectancy (hence the \(^s\) on everything):
+We use the superscript $s$ to index subpopulations comprising the total population, and the superscript ``$1$'' to typically indicate the population with the lower value of $e_0$, and ``$2$'' for the higher value, such that their difference, $\Delta = e_0^2 - e_0^1$, is always positive.
+Causes of death may be indicated with a $c$ subscript, i.e.
+$m_{a,c}$.
+Throughout the manuscript, we assume single-year age groups and omit age interval (left-side) subscripts ($n$) where they would otherwise be due.
 
-\begin{align}
-{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
- &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
-\end{align}
+\subsection*{Averaging life expectancy}\label{averaging-life-expectancy}
 
-\noindent where \(\omega\) is the highest (open) age group, and
-\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
-expectancy being decomposed, and it is composed of age-specific
-contributions, \({}_n\Delta_x^s\), which can be calculated following
-Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll
-use a lifetable radix of 1, meaning \(\ell_0 = 1\) to reduce the formula
-a bit.
+\citet{vaupel2002life} distinguishes between two major approaches to calculate life expectancy for a total population composed of observed subgroups.
+The dominant approach is based on averaging the mortality rate in each age by aggregating deaths and exposures over all subpopulations and calculating the rate per equation~\eqref{eq_rates}:
+
+\begin{equation}
+\label{eq_rates}
+m_a = \frac{\sum_i{D_a^i}}{\sum_i{E_a^i}} = \sum_i \pi_a^i m_a^i \quad \mathrm{,}
+\end{equation}
+
+\noindent where $D$ and $E$ stand for deaths and exposures, respectively.
+This is the standard method used implicitly or explicitly by national statistical offices, the Human Mortality Database \citep{wilmoth2021methods}, or the World Population Prospects \citep{wpp2022mp}, among others.
+\citet{vaupel2002life} called this approach the \emph{current rates} perspective, which treats a heterogeneous population as a homogeneous one.
+
+A second approach derives independent subgroup-specific lifetables and then weights life expectancies together based on an initial mixing composition to obtain the total life expectancy per equation~\eqref{eq_avg}:
+
+\begin{equation}
+\label{eq_avg}
+e_0 = \sum_s \pi_0^s \cdot e_0^s \quad \textrm{.}
+\end{equation}
+
+\noindent This method aligns with Vaupel's \emph{current conditions} approach, at least to the extent that heterogeneous conditions are accounted for using observed strata.
+This second approach is commonly applied in multistate models with strata combined according to mixed initial conditions to obtain the total population \citep{caswell2018age, caswell2021healthy}, occasionally also with standard lifetables \citep{gupta1988note, muszynska2022mortality}, or when calculating between-within decompositions of variance or other similar summary measures \citep{seaman2019changing, lifeineq}.
+If the composition is more favorable in younger than in older ages (for example when younger generations have higher average educational attainment), the life expectancy obtained with \eqref{eq_avg} will be higher than one based on \eqref{eq_rates}.
+This is the total life expectancy approach that our proposed method is designed to decompose.
+
+\subsection*{Kitagawa decomposition}\label{kitagawa-decomposition}
+
+Equation \eqref{eq_avg} treats the total life expectancy as a weighted average, allowing us to precisely decompose a difference in $e_0$, i.e.
+where $\Delta = e_0^2 - e_0^1$, using the formulas from \citet{kitagawa1955components}.
+We presume that population 2 has the higher life expectancy of the two, such that $\Delta$ is positive.
+Specifically, equation \eqref{eq_kit_comp} gives an overall effect of differences in composition:
+
+\begin{equation}
+\label{eq_kit_comp}
+\Delta^{(\pi)} = \sum_s \left(\pi^{s,2}- \pi^{s,1}\right) \cdot \overline{e_0^s}\quad \textrm{,}
+\end{equation}
+ where 
+ \begin{equation}\label{eq:e0mean}
+     \overline{e_0^s} = \frac{e_0^{s,1} + e_0^{s,2}}{2} 
+ \end{equation}
+
+\noindent This result is widely known.
+Note that the the composition effect must be summed this way because group-specific composition (structure) effects are not well-identified and therefore cannot be interpreted in isolation.
+Equation \eqref{eq_kit_e0} gives the subgroup-specific effects of differences in life expectancy:
+
+\begin{equation}
+\label{eq_kit_e0}
+\Delta^{(e_0^s)} = \left(e_0^{s,2} - e_0^{s,1}\right) \cdot \overline{\pi^s}\quad \textrm{,}
+\end{equation}
+where $\overline{\pi^s}$ is the average composition, analogous to equation~\eqref{eq:e0mean}, and where the superscripts are consistent with those used to calculate the total difference $\Delta$.
+
+Equation~\eqref{eq_kit_sum} states that the observed difference in total life expectancy is the sum of (i) a single component capturing the effect of compositional change and (ii) a set of components giving the contribution of each subgroup's life expectancy difference to the total life expectancy difference.
+This second effect could equivalently be called the \textit{rate} effect because each life expectancy is ultimately a function of mortality rates.
+
+\begin{equation}
+\label{eq_kit_sum}
+\Delta = \Delta^{(\pi)}  + \sum_s \Delta^{(e_0^s)}\textrm{.}
+\end{equation}
+
+\subsection*{Symmetrical Arriaga
+decomposition}\label{symmetrical-arriaga-decomposition}
+
+The \emph{rate} effect as per \eqref{eq_kit_e0} can be viewed as the \emph{net} effect on total life expectancy differences resulting from the age-specific (potentially also cause-specific) differences between like subgroups, as isolated by various life expectancy decomposition methods.
+In this setting, the choice of method to derive age-specific effects for subgroup-specific changes in life expectancy is not crucial.
+The \citet{arriaga1984measuring} decomposition approach is widely favored because it is designed for discrete data, it is analytic (implying computational efficiency), and because decomposition results sum exactly to the observed life expectancy difference.
+Since our application concerns sex differences rather than changes over time, we consider a \emph{symmetrical} Arriaga decomposition.
+Since we also include cause-of-death information, we transform the symmetrical decomposition results into an implied sensitivity function.
+This exercise is repeated for each subgroup to decompose group-specific changes in life expectancy (indicated with the superscript $^s$ on each variable):
+
+\begin{equation}
+\label{eq_kit_e0_s}
+\Delta^s &= e_0^{s,2} - e_0^{s,1} 
+\end{equation}
+
+\noindent In this equation $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, which is composed of age-specific contributions, $\overrightarrow{\Delta_a^s}$ or $\overleftarrow{\Delta_a^s}$ depending on whether we decompose from population 1 to 2 (forward) or from 2 to 1 (backward).
+The forward age-specific values $\overrightarrow{\Delta_a^s}$ can be calculated following Arriaga's decomposition method, consistent with \citet{arriaga1984measuring} or the presentation in \citet{preston2000demography}, as outlined in equation \eqref{eq_arriaga1}.
+We use a lifetable radix of 1, meaning $\ell_0 = 1$, to simplify the formula slightly.
 
@@
 \label{eq_arriaga1}
- {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
-  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
+ \overrightarrow{\Delta_a^s} = \begin{cases}\ell_a^{s,1} \cdot \left( \frac{L_a^{s,2}}{\ell_a^{s,2}} - \frac{L_a^{s,1}}{\ell_a^{s,1}}\right)+T_{a+n}^{s,2}\cdot\left(\frac{\ell_a^{s,1}}{\ell_a^{s,2}} - \frac{\ell_{a+n}^{s,1}}{\ell_{a+n}^{s,2}}\right)  & \text{$ \forall a < \omega$}\textrm{,}  \\
+  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall a = \omega $}\textrm{.}
   \end{cases}
@@
 
-Equation \eqref{eq_arriaga1} is the first pass of our symmetrical
-decomposition, where we move \emph{forward} in time from \(t=1\) to
-\(t=2\), so please permit . It
+Equation \eqref{eq_arriaga1} represents the first pass of our symmetrical decomposition, while \eqref{eq_arriaga2} is the second pass, which is identical except for swapping population superscripts.
+
+\begin{equation}
+\label{eq_arriaga2}
+ \overleftarrow{\Delta_a^s} = \begin{cases}\ell_a^{s,2} \cdot \left( \frac{L_a^{s,1}}{\ell_a^{s,1}} - \frac{L_a^{s,2}}{\ell_a^{s,2}}\right)+T_{a+n}^{s,1}\cdot\left(\frac{\ell_a^{s,2}}{\ell_a^{s,1}} - \frac{\ell_{a+n}^{s,2}}{\ell_{a+n}^{s,1}}\right)  & \text{$ \forall a < \omega$}\textrm{,}  \\
+  \ell_\omega ^{s,2} \cdot (e_\omega^{s,1} -e_\omega^{s,2} ) & \text{$\forall a = \omega $}\textrm{.}
+  \end{cases}
+\end{equation}
+
+Importantly,
+
+\begin{align}
+\label{eq_sum2}
+ -\Delta^s &= \sum_a \overleftarrow{\Delta_a^s} \\
+ &= e_0^1 - e_0^2 \nonumber\textrm{.}
+\end{align}
+
+A symmetrical estimate of $\Delta_a^s$ is given by the sign-adjusted average of \eqref{eq_arriaga1} and \eqref{eq_arriaga2} for each age, again with a preference for a positive difference:
+
+\begin{equation}
+\label{eq_arriaga3}
+\Delta_a^s = \frac{\left(\overrightarrow{\Delta_a^s} -  \overleftarrow{\Delta_a^s}\right)}{2}\textrm{.}
+\end{equation}
+
+Repeat the symmetrical Arriaga steps to derive age-specific contributions, $\Delta_a^s$, for each life expectancy difference between like-defined subgroups.
+To take into account information on causes of death, it is best to derive an ad-hoc all-cause sensitivity function, $s_a^s$, by dividing the decomposition result by the mortality rate difference:
+
+\begin{equation}
+\label{eq:arr_sen}
+s_a^s = \frac{\Delta_a^s}{m_{a}^{s,2} - m_{a}^{s,1}} \quad\mathrm{,}
+\end{equation}
+
+\noindent and then multiply age-cause-specific mortality rate differences into $s_a^s$ to obtain the subgroup-specific decomposition by age and cause per equation~\eqref{eq:arr_mxc}.
+
+\begin{equation}
+\label{eq:arr_mxc}
+\Delta_{a,c}^s = s_a^s \cdot \left( m_{a,c}^{s,2} - m_{a,c}^{s,1}\right)\textrm{.}
+\end{equation}
+
+This way of accounting for causes of death is exactly additive.
+However, if any all-cause mortality rate difference is close to 0, equation~\eqref{eq:arr_sen} is vulnerable, in which case we advise deriving the sensitivity using a more robust (but less precise) approach \citep{coddecomp}
+
+\subsection*{Rescale Arriaga results}\label{rescale-arriaga-results}
+
+ To obtain the \emph{net} impact of $\Delta_a^s$ (or $\Delta_{a,c}^s$) on overall life expectancy change we rescale using \eqref{eq_rescale} to match the life expectancy components from equation \eqref{eq_kit_e0}.
+
+\begin{equation}
+\label{eq_rescale}
+\Delta_{a,c}^{s,\text{net}} = \Delta^{e_0^s} \cdot \frac{\Delta_{a,c}^s}{\Delta^s} \quad \mathrm{,}
+\end{equation}
+
+With the composition effect from equation~\eqref{eq_kit_comp} ( $\Delta^\pi$) and \eqref{eq_rescale}, we have all elements of the proposed decomposition:
+
+\begin{equation}
+\label{eq_full}
+\Delta = \Delta^\pi + \sum_s \sum_a \sum_c\Delta_{a,c}^{s,\mathrm{net}}\mathrm{.}
+\end{equation}
+
+\subsection*{Application}\label{rescale-arriaga-results}
+
+We use data from \citet{trias2023age} on sex-, age-, education-, and cause-specific death counts for individuals aged 35 and over for the years 2016-21 in Spain, which were obtained by request from the Spanish National Statistics Institute (INE).
+We use all-cause mortality rates for each education- and sex-specific subpopulation from the same source.
+Educational attainment information is not recorded on death certificates, but INE adds this variable through multiple data linkages \citet{ine2020asignacion}, including municipal population registers (``Padrón'') and the 2011 census.
+We redistributed deaths with missing education codes ($<2$\% of deaths) across the four educational groups within age, sex, and cause of death classes, proportional to deaths with non-missing education.
+We then grouped educational attainment into three categories: Low (primary education or less), Medium (secondary education), and High (tertiary or university education).
+We grouped causes of death from the original ICD10 4-digit codes into 17 main causes: Infectious (ICD-10 codes: A00-B99), Neoplasms (C00-D48), Blood (D50-D89), Endocrine (E00-E90), Mental (F00-F99), Nervous (G00-G99), Circulatory (00-I99), Respiratory (J00-J99), Digestive (K00-K93), Skin (L00-L99), Musculoskeletal (M00-M99), Genitourinary (N00-N99), Congenital (Q00-Q99), Ill-defined (R00-R94), External (chapters S, T, V, W, X and Y), Other causes (chapters H, O and P), and COVID-19 (U071-U072).
+Moreover, to separate the pandemic period we combined years into two time periods: 2016-2019 and 2020+.
+Our main results refer to 2016-2019, whereas extra results for 2020-2021 are given in the materials repository.
+We smoothed and graduated the original data from 5-year age groups to single ages using a generalized additive model (GAM), including a P-spline over age and population offsets \citep{mgcv2011,mgcv2017}.
+The GAM formula is given in equation~\eqref{gam_model}, using the Quasi-Poisson family of models to account for possible overdispersion or Negative Binomial model in cases with a more substantial overdispersion:
+
+\begin{equation}
+\label{gam_model}
+\log(\mu_a) = \beta_0 + f(a) + ln(pop_a)
+\textrm{.}
+\end{equation}
+
+\FloatBarrier
+\section*{Results}\label{results}
+We present empirical results for the decomposition of the sex difference in life expectancy in 2016-2019.
+Table 1 presents the components of the Kitagawa decomposition for males and females.
+In this context,  $\pi$ c represents the educational structure, while $e(35)$ denotes life expectancy at age 35 (``rate'' element in original Kitagawa terms).
+The table also gives the corresponding education structure and life expectancy differences between sexes.
+The final element,  $\Delta^{(e_{35}^s)}$, is the decomposition result, which gives the education group contribution.
+The largest life expectancy gap is observed among individuals with Primary and Secondary education levels, and this gap is lower for those with Higher education.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{l|lccccccc}
+  \hline
+  Education  & $\pi^f$ & $\pi^m$ & $\pi^f - \pi^m$ & $e(35)^f$ & $e(35)^m$ & $\overline{e(35)}$ & \makecell{$e(35)^f -$\\$ e(35)^m$} & $\Delta^{(e_{35}^s)}$ \\ 
+  \hline
+   Primary  & 0.09 & 0.10 & -0.01 & 48.21 & 42.47 & 45.34 & 5.74 & 0.54\\   
+   Secondary  & 0.41 & 0.52 & -0.11 & 50.29 & 44.52 & 47.40 & 5.78 & 2.69\\ 
+   Higher  & 0.50 & 0.38 & 0.12 & 51.30 & 46.92 & 49.11 & 4.38 & 1.92\\   
+   \hline \hline
+\end{tabular}
+\caption{Elements of the Kitagawa decomposition, sex gaps in life expectancy in Spain, 2016-19}
+\label{tab:my_table}
+\end{table}
+
+The corresponding difference in education-specific life expectancy is also illustrated in Figure 1.
+This figure additionally shows the differences observed for standard aggregated (non-stationary) versus radix-weighted life expectancy (stationary).
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=.8\linewidth]{manuscript/fig1.pdf}
+    \caption{The Female-Male difference in $e(35)$ by education and population type, Spain 2016-19. }
+    \label{fig:enter-label}
+\end{figure}
+Figure 2 is complementary to Figure 1 and represents the education contribution to the $e(35)$ sex gap with the educational composition component contribution, about 3 months, being added to the figure.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=.8\linewidth]{manuscript/fig2.pdf}
+    \caption{Education structure contribution to total $e(35)$ sex gap, Spain 2016-19}
+    \label{fig:enter-label}
+\end{figure}
+
+Figure 3 illustrates the cause-specific contributions to the life expectancy gap.
+The majority of this gap is explained by differences in mortality from cancer, circulatory diseases, respiratory conditions, and external causes.
+Neoplasms alone contribute to almost two full years of life expectancy.
+The only categories where males have a slight advantage in mortality are musculoskeletal and other minor causes.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=.8\linewidth]{manuscript/fig3.pdf}
+    \caption{Cause-specific contribution to sex difference in total $e(35)$, Spain 2016-19. The Education structure contribution is presented in a different colour}
+    \label{fig:enter-label}
+\end{figure}
+
+Further decomposition of cause-specific differences for age and education components of composed decomposition is shown in Figure 4.
+The majority of the difference in neoplasm mortality is concentrated in the older age group of 70-80 years.
+The circulatory component is observed in slightly younger ages and is more uniformly distributed between ages 60 and 80.
+In contrast, the respiratory component is more pronounced around age 85.
+The external causes of death primarily contribute to the difference in younger ages and since our data is bounded below by the age of 35, the full effect cannot be fully observed here.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=0.8\linewidth]{manuscript/fig4.pdf}
+    \caption{The contributions of age, education, and major causes of death to the sex gap in total $e(35)$.}
+    \label{fig:enter-label}
+\end{figure}
+
+\FloatBarrier
+
+\section*{Discussion}\label{discussion}
+
+In this paper, we propose a method for decomposing the differences in life expectancy that accounts for the compositional effects of education and causes of death between two populations mixed in the radix.
+Deriving an analytical solution for decomposition offers several advantages over alternative methods, particularly in terms of simplicity, speed, and repeatability.
+Unlike the linear integral method \citet{horiuchi2008decomposition}, which requires programming expertise, our approach can be executed swiftly and efficiently, even within spreadsheet-like environments.
+The computational efficiency of our method allows for the calculation of confidence intervals using bootstrapping.
+
+Using the Kitagawa decomposition approach \citep{kitagawa1955components} to weigh together group-specific paired decompositions, our method can be generalized to address many decomposition problems involving structural compositional components, and it is not bound to mortality analyses.
+However, this post-weighting approach is most suitable for cases where populations are blended in a radix.
+In contrast, other approaches in the literature  \citep{torres2019contribution,su2023educational,su2024subnational,hendi2021immigration} decompose referring to the standard \emph{current rates} \citep{vaupel2002life} method of calculating life expectancy.
+These approaches treat the age pattern of group prevalence differently, in essence fixing group prevalence (weights) rather than making prevalence depend on mortality.
+We point out that mortality is often one of the major drivers of how group prevalence weights change over age.
+
+ When incorporating information on causes of death, a potential vulnerability of our method arises when the difference in mortality rates in the denominator of equation~\eqref{eq:arr_sen} approaches zero (compare with Box 4.3 of \citet{preston2000demography}), rendering an implausible result.
+In this case, one may prefer to use a direct approximation or numerical estimate of the life expectancy sensitivity in equation~\eqref{eq:arr_mxc}.
+
+Our empirical results indicate that the sex gap in life expectancy remains a significant issue in Spain across all educational strata.
+However, it is about 25\% lower in the Higher education group.
+This can be attributed to several factors, including females' lower engagement in risky behaviours \citep{ross2012education,cook2001knowing,byrnes1999gender,kritsotakis2016gender,olson2017gender}, better jobs and greater health awareness of males with higher education \citep{lawrence2017college,ross1995links,ross2012education, mcmahon2009higher}.
+In terms of causes, neoplasms, circulatory diseases, and respiratory diseases account for the largest contributions to the sex gap.
+Given that the male disadvantage peaks around ages 65-85 (birth cohorts 1934-1951), this difference can be partly explained by the differences in smoking patterns between males and females within each subpopulation \citep{haeberer2020social}.
+Since our findings are based on education as a marker of socioeconomic status, they refer only to ages above 35.
+Therefore, we do not here measure the full power of socioeconomic status or different causes of death in driving the overall sex gap in life expectancy.
+Further studies could use our method to explain the sex gap using geography, socioeconomic status, and causes of death.
+
+\hypertarget{Conclusion}{%
+\section*{Conclusion}\label{conclusion}
+
+Our proposed combination of two analytical decomposition methods can handle the problem of a composed life expectancy, thereby giving an efficient framework for analyzing differences in life expectancy while accounting for population heterogeneity on fixed characteristics.
+We give both R code and spreadsheet implementations of the method.
+Our framework is straightforward to use and does not require high computational capacity, making it suitable for implementation in a spreadsheet-like environment while providing results comparable to those of the widely used Horiuchi method.
+The main advantages of the proposed method are its computational efficiency and ease of implementation over the commonly used Horiuchi method.
+Further efforts in alleviating the sex gap in mortality are required.
+The main cause of death contributions suggests that attention to the determinants of neoplasms, cardiovascular, and respiratory diseases could play a substantial role in reducing the sex gap in life expectancy.
+
+\bibliography{bibliography.bib}
 
@@
 
-If your article has accompanying supplementary file/s please state so
-here.
+If your article has accompanying supplementary file/s please state so here.
 
@@
 
-Acknowledgments are not compulsory. Where included they should be brief.
-Grant or contribution numbers may be acknowledged.
+Thanks to Michael Lachanski for posing the question that led to the solution we present here, to Jonas Schoeley for first raising our attention to the composition problem with the Kitagawa method, and to the MPIDR Laboratory of Population Health for posing assorted questions in September 2023 that also led to this work.
 
-Please refer to Journal-level guidance for any specific requirements.
+\subsection*{Funding}\label{funding}
 
-\hypertarget{declarations}{%
-\section*{Declarations}\label{declarations}}
-\addcontentsline{toc}{section}{Declarations}
+This work was partly financed by grant nr SR22-00502 from \emph{la Caixa Foundation} and grant number PID2022-142762OA-I00 from the Spanish \emph{Ministerio de Ciencia e Innovation}.
+STL acknowledges research funding from the Ramon y Cajal program of the same Ministry (RYC2021-033123-I).
 
-Some journals require declarations to be submitted in a standardised
-format. Please check the Instructions for Authors of the journal to
-which you are submitting to see if you need to complete this section. If
-yes, your manuscript must contain the following sections under the
-heading `Declarations':
+\subsection*{Conflict of interest}\label{conflict-of-interest}
 
-\begin{itemize}
-\tightlist
-\item
-  Funding
-\item
-  Conflict of interest/Competing interests (check journal-specific
-  guidelines for which heading to use)
-\item
-  Ethics approval
-\item
-  Consent to participate
-\item
-  Consent for publication
-\item
-  Availability of data and materials
-\item
-  Code availability
-\item
-  Authors' contributions
-\end{itemize}
+The authors declare no conflict of interest
 
-\noindent If any of the sections are not relevant to your manuscript,
-please include the heading and write `Not applicable' for that section.
+\subsection*{Availability of data and
+materials}\label{availability-of-data-and-materials}
 
-\begin{appendices}
+We cannot share the original data that this study is based on. We do
 
-\hypertarget{secA1}{%
-\section{Section title of first appendix}\label{secA1}}
+\subsection*{Code availability}\label{code-availability}
 
-An appendix contains supplementary information that is not an essential
-part of the text itself but which may be helpful in providing a more
-comprehensive understanding of the research problem or it is information
-that is too cumbersome to be included in the body of the paper.
-
-For submissions to Nature Portfolio Journals please use the heading
-``Extended Data''.
-
-\end{appendices}
-
-\bibliography{bibliography.bib}
-
+All code used here is available in an open \texttt{GitHub} repository
+\url{https://github.com/timriffe/arriagakitagawa}
+}
 \end{document}

### 2024-09-12 R/01_smoothing_and_ungroupping.R
adapts code for partial reproducibility; reorders figures in fig script

@@ -47,3 +47,6 @@ mxc_single <- data_5_prepped |>
                       smooth_ungroup(.x, new_data = new_data)
   )) |> 
   unnest(data)
+
+
+write_csv(mxc_single,"data/mxc_single.csv")

### 2024-09-12 R/02_LT_and_average_quantities.R

@@ -1,6 +1,14 @@
-source("R/00_initial_data_preparation.R")
-source("R/01_smoothing_and_ungroupping.R")
+# source("R/00_initial_data_preparation.R")
+# source("R/01_smoothing_and_ungroupping.R")
+library(tidyverse)
+library(coddecomp)
+library(ggridges)
+library(xtable)
+library(scales)
+library(ggpubr)
+library(mgcv)
 
+mxc_single <- read_csv("data/mxc_single.csv")
 # LT and basic quantities
 Lt <- mxc_single |>
   filter(cause == "All") |> 

### 2024-09-12 R/03_decomposition_results.R

@@ -1,5 +1,5 @@
-source("R/00_initial_data_preparation.R")
-source("R/01_smoothing_and_ungroupping.R")
+# source("R/00_initial_data_preparation.R")
+# source("R/01_smoothing_and_ungroupping.R")
 source("R/02_LT_and_average_quantities.R")
 
 mxc_decomp <- mxc_single |>
@@ -25,29 +25,40 @@ e35_kit <- Lt |>
          e35_avg = (e35_Females + e35_Males) / 2)
 # ----------------------------------------------------------------------- #
 # population prevalence data from original source
-struct_kit <- data_5_prepped |> 
-  as_tibble() |>
-  filter(year  != "Total",
-         sex   != "Total",
-         educ  !="Total",
-         cause == "All",
-         age   == 35) |> 
-  group_by(sex, educ, year) |> 
-  summarise(deaths  = sum(deaths),
-            pop     = sum(pop),
-            .groups = "drop") |> 
-  group_by(sex, educ, year) |> 
-  # population by groups
-  summarise(pop = sum(pop), .groups = "drop") |> 
-  group_by(sex, year) |>
-  # scale population
-  mutate(prev = pop / sum(pop)) |> 
-  ungroup() |> 
-  select(-pop) |> 
-  group_by(educ, year)  |> 
-  pivot_wider(names_from   = sex, 
-              values_from  = prev, 
-              names_prefix = "st_") |> 
+if (exists(data_5_prepped)){
+  prev <- data_5_prepped |> 
+    as_tibble() |>
+    filter(year  != "Total",
+           sex   != "Total",
+           educ  !="Total",
+           cause == "All",
+           age   == 35) |> 
+    group_by(sex, educ, year) |> 
+    summarise(deaths  = sum(deaths),
+              pop     = sum(pop),
+              .groups = "drop") |> 
+    group_by(sex, educ, year) |> 
+    # population by groups
+    summarise(pop = sum(pop), .groups = "drop") |> 
+    group_by(sex, year) |>
+    # scale population
+    mutate(prev = pop / sum(pop)) |> 
+    ungroup() |> 
+    select(-pop) |> 
+    group_by(educ, year)  |> 
+    pivot_wider(names_from   = sex, 
+                values_from  = prev, 
+                names_prefix = "st_") 
+
+  prev |> 
+    write_csv("data/prev.csv")
+} else {
+  # otherwise read in the last one saved (also committed)
+  prev <- read_csv("data/prev.csv")
+}
+
[…]

### 2024-09-12 R/04_gaps.R

@@ -1,5 +1,5 @@
-source("R/00_initial_data_preparation.R")
-source("R/01_smoothing_and_ungroupping.R")
+# source("R/00_initial_data_preparation.R")
+# source("R/01_smoothing_and_ungroupping.R")
 source("R/02_LT_and_average_quantities.R")
 source("R/03_decomposition_results.R")
 # gaps

### 2024-09-12 R/05_figures.R

@@ -1,43 +1,8 @@
-source("R/00_initial_data_preparation.R")
-source("R/01_smoothing_and_ungroupping.R")
+# source("R/00_initial_data_preparation.R")
+# source("R/01_smoothing_and_ungroupping.R")
 source("R/02_LT_and_average_quantities.R")
 source("R/03_decomposition_results.R")
 source("R/04_gaps.R")
-
-# figure 3
-decomp_total |>
-  filter(year == "2016-2019") |>
-  group_by(cause) |>
-  summarize(margin = sum(result_rescaled)) |>
-  filter(cause != "Covid-19") |>
-  full_join(dplyr::select(st_gaps[1, ], margin = cc_str) |>
-              mutate(cause = "Educ. component")) |>
-  mutate(margin_sign = if_else(sign(margin) == 1, "#F8766D", "#C77CFF")) |>
-  mutate(margin_sign = if_else(cause == "Educ. component", "#00BFC4", margin_sign)) |>
-  ggplot(aes(
-    y = reorder(cause, margin),
-    x = margin,
-    color = margin_sign,
-    fill = margin_sign
-  )) +
-  geom_col() +
-  guides(color = "none") +
-  theme_minimal() +
-  scale_color_identity() +
-  scale_fill_identity() +
-  scale_x_continuous(breaks = pretty_breaks()) +
-  xlab("Contribution to sex-gap (years)") +
-  theme(
-    legend.position = "none",
-    axis.title.y = element_blank(),
-    axis.title.x = element_text(face = "bold", color = "black"),
-    strip.background = element_blank(),
-    strip.text = element_text(face = "bold", color = "black"),
-    legend.text = element_text(face = "bold", color = "black"),
-    legend.title = element_text(face = "bold", color = "black"),
-    axis.text.y = element_text(color = "black", face = "bold"),
-    axis.text.x = element_text(color = "black")
-  )
 
 # figure 1
 education |>
@@ -85,6 +50,76 @@ education |>
   ) +
   labs(fill = "Education groups", ) +
   labs(y = "Difference (years)")
+# ----------------------------------------------------- #
+
+# figure 2
+decomp_total |>
+  filter(year == "2016-2019") |>
+  group_by(educ) |>
+  summarize(margin = sum(result_rescaled)) |>
+  full_join(dplyr::select(st_gaps[1, ], margin = cc_str) |>
+              mutate(educ = "Educ. component")) |>
+  mutate(educ = factor(
+    educ,
[…]

### 2024-09-12 R/06_figures_post_covid.R
re-order post covid figs

@@ -1,42 +1,8 @@
-source("R/00_initial_data_preparation.R")
-source("R/01_smoothing_and_ungroupping.R")
+# source("R/00_initial_data_preparation.R")
+# source("R/01_smoothing_and_ungroupping.R")
 source("R/02_LT_and_average_quantities.R")
 source("R/03_decomposition_results.R")
 source("R/04_gaps.R")
-
-# figure 3 post covid
-decomp_total |>
-  filter(year != "2016-2019") |>
-  group_by(cause) |>
-  summarize(margin = sum(result_rescaled)) |>
-  filter(cause != "Covid-19") |>
-  full_join(dplyr::select(st_gaps[1, ], margin = cc_str) %>%
-              mutate(cause = "Educ. component")) %>%
-  mutate(margin_sign = if_else(sign(margin) == 1, "#F8766D", "#C77CFF")) |>
-  mutate(margin_sign = if_else(cause == "Educ. component", "#00BFC4", margin_sign)) %>%
-  ggplot(aes(
-    y = reorder(cause, margin),
-    x = margin,
-    color = margin_sign,
-    fill = margin_sign
-  )) +
-  geom_col() +
-  guides(color = "none") +
-  theme_minimal() +
-  scale_color_identity() +
-  scale_fill_identity() +
-  scale_x_continuous(breaks = pretty_breaks()) +
-  xlab("Contribution to sex-gap") +
-  theme(
-    legend.position = "none",
-    axis.title = element_blank(),
-    strip.background = element_blank(),
-    strip.text = element_text(face = "bold", color = "black"),
-    legend.text = element_text(face = "bold", color = "black"),
-    legend.title = element_text(face = "bold", color = "black"),
-    axis.text.y = element_text(color = "black", face = "bold"),
-    axis.text.x = element_text(color = "black")
-  )
 
 # figure 1 post covid
 education |>
@@ -84,6 +50,75 @@ education |>
   ) +
   labs(fill = "Education groups", ) +
   labs(y = "Difference (years)")
+# -------------------------------------------- #
+
+# figure 2 post covid
+decomp_total |>
+  filter(year != "2016-2019") |>
+  group_by(educ) |>
+  summarize(margin = sum(result_rescaled)) |>
+  full_join(dplyr::select(st_gaps[1, ], margin = cc_str) |>
+              mutate(educ = "Educ. component")) |>
+  mutate(educ = factor(
+    educ,
+    levels = c("Educ. component", "Higher", "Secondary", "Primary")
[…]
